Cis-regulatory modes of Ultrabithorax inactivation in butterfly forewings

Version 1

Abstract

Abstract

Hox gene clusters encode transcription factors that drive regional specialization during animal development: e.g. the Hox factor Ubx is expressed in the insect metathoracic (T3) wing appendages and differentiates them from T2 mesothoracic identities. Hox transcriptional regulation requires silencing activities that prevent spurious activation and regulatory crosstalks in the wrong tissues, but this has seldom been studied in insects other than Drosophila, which shows a derived Hox dislocation into two genomic clusters that disjoined Antennapedia (Antp) and Ultrabithorax (Ubx). Here we investigated how Ubx is restricted to the hindwing in butterflies, amidst a contiguous Hox cluster. By analysing Hi-C and ATAC-seq data in the butterfly Junonia coenia, we show that a Topologically Associated Domain (TAD) maintains a hindwing-enriched profile of chromatin opening around Ubx. This TAD is bordered by a Boundary Element (BE) that separates it from a region of joined wing activity around the Antp locus. CRISPR mutational perturbation of this BE releases ectopic Ubx expression in forewings, inducing homeotic clones with hindwing identities. Further mutational interrogation of two non-coding RNA encoding regions and one putative cis-regulatory module within the Ubx TAD cause rare homeotic transformations in both directions, indicating the presence of both activating and repressing chromatin features. We also describe a series of spontaneous forewing homeotic phenotypes obtained in Heliconius butterflies, and discuss their possible mutational basis. By leveraging the extensive wing specialization found in butterflies, our initial exploration of Ubx regulation demonstrates the existence of silencing and insulating sequences that prevent its spurious expression in forewings.

Introduction

Hox genes are key specifiers of antero-posterior regional identity in animals, and thus require robust regulatory mechanisms that confine their expression to well-delimited sections of the body. Their genomic arrangement into Hox gene clusters has provided a rich template for the study of gene regulation, with mechanisms including chromatin silencing and opening, 3D conformational changes, and non-coding RNAs (Mallo and Alonso 2013). However, this rich body of work has been almost exclusively performed in mice and fruit flies. In order to decipher how diverse body plans and morphologies evolved, we must start assessing the mechanisms of Hox gene regulation in a wider range of organisms.

The Ultrabithorax (Ubx) gene encodes a Hox family transcription factor involved in the specification of segment identities in arthropods (Hughes and Kaufman 2002; Heffer and Pick 2013). In insects, the conserved expression of Ubx in the metathoracic (T3) segment is required for their differentiation from Ubx-free tissues in the mesothorax (T2), and has been a key factor for the specialization of metathoracic serial appendages including T3 legs (Mahfooz et al. 2007; Refki et al. 2014; Tomoyasu 2017; Feng et al. 2022; Buffry et al. 2023) and hindwings or their derivatives (Tomoyasu 2017; Loker et al. 2021). The mechanisms of Ubx segment-specific expression have been intensively studied in D. melanogaster (Mallo and Alonso 2013; Hajirnis and Mishra 2021), where Hox genes are separated into two genomic loci, the Antennapedia (ANT-C, Antp) and Bithorax clusters (BX-C). In short, the BX-C complex that includes Ubx, abdominal-A (abd-A), and Abdominal-B (Abd-B) is compartmentalized into nine chromosomal domains that determine the parasegmental expression boundaries of these three genes (Maeda and Karch 2015). Each boundary is primarily enforced by insulators that separate Topologically Associating Domains (TADs) of open-chromatin, while also allowing interactions of enhancers with their target promoters (Postika et al. 2018; Srinivasan and Mishra 2020). The BX-C locus also includes non-coding RNAs, some of which are processed into miRNAs known to repress abd-A and Abd-B (Garaulet and Lai 2015). Fub-1/bxd long non-coding RNAs (lncRNAs) situated 5’ of Ubx are thought to participate in Ubx regulation in the PS5 (posterior T3 to anterior A1) parasegment (Ibragimov et al. 2022). An intronic lncRNA dubbed lncRNA:PS4 is expressed in the PS4 parasegment (posterior T2 - anterior T3), and appears to stabilize Ubx in this region in mutant contexts (Hermann et al. 2022). Little is known about how Hox genes are regulated outside of flies, where they co-localize into a single Hox cluster, and where Antp and Ubx thus occur in contiguous positions (Gaunt 2022; Mulhair and Holland 2022). A few Hox-related miRNAs are evolutionarily conserved across the locus in arthropods (Pace et al. 2016), and an early study in Tribolium characterized the embryonic expression of a Hox cluster non-coding transcripts (Shippy et al. 2008).

These knowledge gaps lead us to consider the use of butterflies and moths (Lepidoptera) as alternative model systems for the study of Ubx function and regulation. Lepidopteran forewings and hindwings are functionally and morphologically differentiated, and CRISPR mosaic knock-outs (mKOs) showed that Ubx is necessary for the specification of hindwing color patterns, shape, and venation (Tendolkar et al. 2021). In three species of nymphalid butterflies (Heliconius erato, Junonia coenia, and Bicyclus anynana), CRISPR-mediated loss-of-function of Ubx induces regional-specific homeotic transformations of hindwing patterns into their forewing counterpart (Matsuoka and Monteiro 2018; Tendolkar et al. 2021), reminiscent of homeotic aberrations that are sporadically observed in butterfly wings (Sibatani 1983; Nijhout and Rountree 1995). The ectopic activation of Ubx into the pupal forewing results in the gain of hindwing features, suggesting Ubx is sufficient to drive T3-like identity when expressed in T2 (Lewis et al. 1999; Tong et al. 2014). Besides its roles in adult wing differentiation, Ubx is also known to repress thoracic leg identity in transient embryonic appendages of the first abdominal segment, called pleuropods (Zheng et al. 1999; Masumoto et al. 2009; Tong et al. 2017; Tendolkar et al. 2021; Matsuoka et al. 2022).

The general organization of Hox gene clusters has been well described in Lepidoptera, but their regulation has been seldom studied. Lepidopteran genomes have accumulated divergent Hox3 copies, named Shox genes, that are required during early embryonic development but do not appear to play homeotic functions (Ferguson et al. 2014; Livraghi 2017; Mulhair et al. 2022). An lncRNA and two miRNAs were identified in the intergenic region between abd-A and Abd-B in the silkworm (Wang et al. 2017, 2019). In butterfly wings, the regulation of Ubx shows strong patterns of segment-specific regulation at two levels. First, the Ubx transcript has been consistently identified as the most differentially expressed mRNA between the two wing sets (Hanly et al. 2019; Wang et al. 2022). Second, comparison of ATAC-seq signals reveal that forewing vs. hindwing have identical open-chromatin profiles during development across the genome, except at the Ubx gene itself (Lewis and Reed 2018; van der Burg et al. 2019). Thus, the ability of the Ubx locus to be robustly activated in hindwings and repressed in forewings is likely driving most subsequent differences between these tissues. In this study, we provide an initial assessment of the regulation of the Ubx locus during butterfly wing development. To do this, we leverage genomic resources and CRISPR mutagenesis with a focus on two laboratory species belonging to the Nymphalinae sub-family, J. coenia and Vanessa cardui (Livraghi et al. 2017; Martin et al. 2020; van der Burg et al. 2020; Mazo-Vargas et al. 2022). We identify putative regulatory regions likely involved in the repression and activation of Ubx expression, and discuss the potential mechanisms restricting it to hindwings. Finally, we describe a collection of spontaneous wing homeotic mutants in Heliconius spp. and elaborate on the categories of mutations that could underlie these phenotypes by misregulating Ubx.

Results

Gene expression at the Ubx locus during wing development

We provide annotations of the Ubx genomic region in four Nymphalinae butterflies (Fig. 1A). These feature existing genomic resources for our model species J. coenia and V. cardui (van der Burg et al. 2020; Lohse et al. 2021b; Zhang et al. 2021), as well as for Aglais (Nymphalis) io (Lohse et al. 2021a). The publicly available annotations for these three species include evidence from developmental transcriptomes, and we added to this set a manual annotation of the Ubx locus from the oak leaf butterfly Kallima inachus, for which forewing vs. hindwing transcriptomes have been sequenced across a replicated developmental time series (Yang et al. 2020; Wang et al. 2022).

Annotation of the Ubx genomic interval in four butterflies of the Nymphalinae sub-family.

(A) Genomic intervals spanning Antp, Ubx, and abd-A, featuring published transcript annotations from NCBI Reference Genomes of V. cardui and A. io, and manual re-annotations of the J. coenia and K. inachus genomes using published RNAseq dataset (see Methods). Exons are shown with coding (thick) and non-coding (thin) sections. No lincRNA:Ubx-AS5’ transcripts were detected in K. inachus. (B) Expression profiling of transcripts of the Ubx region in K. inachus, based on a reanalysis of published wing RNA-seq transcriptomes (Wang et al. 2022). Expression levels are plotted as DESeq2 normalized counts plots. Pairwise Wald tests adjusted for multiple test correction each assess differential expression between forewings and hindwings. ns : non-significant ; * : p < 0.05; ** : p < 0.01 ; *** : p < 0.001.

All Nymphalinae show a similar organization of the region spanning Ubx. Interestingly, the first intron of Ubx encodes a long non-coding RNA in opposite orientation to Ubx, that we dub here lncRNA:Ubx-IT1 (abbr. Ubx-IT1), based on the recommended nomenclature (Seal et al. 2022).

Orthologous versions of Ubx-IT1 are detected in most NCBI RefSeq genome annotations throughout Lepidoptera (e.g. the ncRNA NCBI:XR_960726 in Plutella xylostella), implying it is a conserved feature of the Ubx locus in this insect order. Finally, both annotations from V. cardui, A. io, and J. coenia show a long intergenic non-coding transcript starting in antisense orientation about 10-15 kb 5’ of Ubx, that we dub here lincRNA:Ubx-AS5’ (abbr. Ubx-AS5’). This transcript was neither detected in K. inachus transcriptomes nor in RNA datasets outside of the Nymphalinae sub-family, and could be specific to this lineage.

Next we reanalyzed the K. inachus wing transcriptomes (Wang et al. 2022), and profiled the expression of Ubx region transcripts during wing development (Fig. 1B). As expected from previous studies (Hanly et al. 2019; Paul et al. 2021; Merabet and Carnesecchi 2022; Wang et al. 2022), Ubx showed a strong expression bias in hindwings, spanning the larval imaginal disks to the intermediate pupal stage. Of note, Ubx is confined to the peripodial membranes of insect T2 wing appendages (Weatherbee et al. 1998, 1999; Prasad et al. 2016), which may explain residual detection in some of the forewing samples here. Ubx-IT1 was significantly enriched in hindwings compared to forewings, albeit at ∼100-fold lower count levels than Ubx in the same samples. The Hox gene Antp showed a minor forewing enrichment, confirming that while Ubx expression is robustly repressed in T2 forewing tissues, Antp expression is permitted in both T2 and T3 appendages (Matsuoka and Monteiro 2021, 2022; Paul et al. 2021). Expression of abd-A was undetected in most wing samples.

Chromatin 3D conformation reveals a Boundary Element between Antp and Ubx

Genome-wide Hi-C sequencing can be used to generate heatmaps that capture the conformation of 3D chromatin in tissues, and has been used extensively to study Drosophila Hox cluster organisation into TADs that prevent regulatory crosstalk between adjacent genes (Ibragimov et al. 2022; Moniot-Perron et al. 2023). Here we used Hi-C to assess the 3D chromatin architecture of the Hox cluster interval in the butterfly J. coenia, using existing datasets that were generated from fifth instar larval forewings (van der Burg et al. 2020; Mazo-Vargas et al. 2022). In larval forewings, the Hox chromatin conformation landscape consists of three well-delimited TADs, the first one spanning proboscipedia (pb) to Sex comb reduced (Scr), the second one around Antp, and the third one Ubx, abd-A, and Abd-B (Figs. 2 and 3A). A Boundary Element (BE), was robustly called (see Methods) at the region separating the Antp and Ubx TADs, situated in the Ubx last intron. Because TAD boundary prediction has a coarse resolution, we arbitrarily define the candidate BE region as a 15-kb interval centered in the Ubx last intron, and dub it Antp-Ubx_BE. A binding motif analysis identified 4 CTCF binding sites in a 1-kb interval within Antp- Ubx_BE, two of which were found in a tightly linked, convergent orientation (Fig. S1), which is consistent with TAD insulating role in mediating chromatin loop-extrusion (Guo et al. 2015). This concordance between Hi-C profiling and CTCF motif prediction thus indicates that Antp-Ubx_BE region functions as an insulator between regulatory domains of Antp and Ubx.

A region of hindwing-specific chromatin-opening is bordered by a TAD BE in the last intron of Ubx.

(A) Hi-C contact heatmap in fifth instar forewings of J. coenia and TAD separation scores around Ubx. A TAD boundary element (Antp-Ubx_BE) is inferred in the last intron of Ubx (vertical dotted line). (B) Differential ATAC-seq profiles, re-analyzed from a previous dataset (Mazo-Vargas et al. 2022). Top : open-chromatin profiles of forewings (FW, magenta), and hindwings (HW, green), respectively subtracted from larval head signal (purple, negative when wing signals at background-level). Bottom : subtractive ATAC-seq profile (HW-FW) revealing hindwing-enriched chromatin in the Ubx locus. Antp-Ubx_BE is in the vicinity of an isolated region of forewing-enriched opening (blue arrowhead). (C) PhastCons genomic alignment scores, with overall alignability suggesting minimal structural variation across this interval in Lepidoptera and Trichoptera.

Hindwing-enriched chromatin-opening around Ubx, and the Antp-Ubx_BE boundary, are both maintained in mid-pupal hindwings.

(A) Hi-C heatmap in J. coenia fifth instar larval forewings, and subtractive ATAC-seq profiles at this stage (hindwing-forewing), as expanded from Fig. 2 across the Hox cluster. (B) Hi-C heatmap in J. coenia mid-pupal hindwings, and subtractive ATAC-seq profiles at this stage (forewing-hindwing). Inferred TAD boundaries are shown as vertical dotted lines. Blue arrowhead : position of the Antp-Ubx_BE sgRNA.

Differential forewing vs. hindwing chromatin-opening across the Antp-Ubx interval

In flies, the Ubx/abd-A section is organized into regulatory domains that display differential activities across the antero-posterior axis, following what has been called the open-for-business model (Maeda and Karch 2015; Gaunt 2022). Here we tested if this pattern extends to butterfly species with a contiguous Hox cluster. To do this we used ATAC-seq datasets from J. coenia forewing (T2), hindwing (T3), and whole-head tissues sampled across fifth instar larval and early pupal stages, similarly to previous studies (van der Burg et al. 2020; Mazo-Vargas et al. 2022; Van Belleghem et al. 2023). These data reveal that chromatin opening along the Antp/Ubx/abd-A interval is partitioned into several regions showing a transition of T2 to T3 activity (Fig. 2B). From the anterior to posterior Hox colinear order (i.e. from Antp towards abd-A), chromatin-opening forms a block of forewing-enriched activity close to Antp and its 5’ region, to a block of activity in both forewings and hindwings that stops at the Antp-Ubx_BE. This region is consistent with the fact that Antp is expressed in both wing pairs (Fig. 1B). From Antp-Ubx_BE, the interval including Ubx and a large upstream region is strongly enriched for hindwing opening, consistently with previous studies that found it to be the only genomic region showing this pattern (Lewis and Reed 2018; van der Burg et al. 2019). Last, the region surrounding abd- A is devoid of differntial open-chromatin activity between forewings and hindwings, in accordance with the exclusion of its expression from thoracic segments (Warren et al. 1994; Tong et al. 2014)

Comparison of 3D conformation and open-chromatin profiles between larval forewings and mid- pupal hindwings

The Hi-C dataset analyzed above was prepared from larval forewings, and forewings do not express Ubx (Fig. 1B). Next, we repeated our analysis on a Hi-C dataset generated in pupal hindwings instead (van der Burg et al. 2020), i.e. in a later-stage tissue expressing Ubx. We found two main differences in this contact landscape compared to the larval forewing (Fig. 3). First, the TAD spanning from proboscipedia (pb) to fushi-tarazu (ftz) faded in intensity, while in contrast, the TAD around Antp remained strongly organized. Second, Ubx lost its physical association to the abd-A and Abd-B domains, and gained a TAD boundary situated in the Ubx-AS5’ intron. It is difficult to disentangle effects from staging (larval vs. pupal) and tissues (forewing vs. hindwing) in this comparison. Specifically, these differences we observed may be due to chromatin remodeling between stages, as somewhat expected during metamorphosis (Gutierrez-Perez et al. 2019). Alternatively, it is also possible hindwing development requires insulate Ubx from more posterior enhancers. These issues will require further investigation, for instance using profiling of histone marks, with pairwise forewing-hindwing comparison at single stages. Nonetheless the later hindwing sample showed a maintenance of Antp-Ubx separation. First, while Ubx formed a smaller TAD spanning its coding exons 1-2, this region conserved a domain of hindwing- enriched open-chromatin. Second, boundary prediction called two possible, tightly linked TAD limits in the Antp-Ubx_BE region, showing that the last intron of Ubx still acts as an insulating region. In conclusion, our preliminary comparison of Hox 3D conformation indicates that the Antp-Ubx_BE is relatively stable across two stages and wing serial homologs.

Mutagenic perturbation of Antp-Ubx_BE results in forewing homeosis

Next, we reasoned that the forewing-enriched ATAC-seq peak present in the inferred boundary interval (Fig. 4A) might mediate the binding of insulator proteins (Savitsky et al. 2016; Stadler et al. 2017), or act as a transcriptional silencer (Segert et al. 2021). Several genomic features support the former hypothesis. First, the only forewing-enriched ATAC-seq peak across a 150-kb region (spanning the Ubx gene and the Antp-Ubx intergenic region), coincides with the midpoint between the two tentative hicFindTADs boundary predictions inferred from HiC data (Fig. 2B). Second, during motif scans conducted across that 150-kb region we found 8 predicted binding-sites for the Drosophila CCCTC- Binding Factor (CTCF) clustered in a 5-kb region around the differentially accessible region, and none elsewhere in the last Ubx intron (Fig. 4A), suggesting the forewing-enriched ATAC-seq peak may function as a transcriptional insulator (Gambetta and Furlong 2018; Postika et al. 2018; Kyrchanova et al. 2020; Kaushal et al. 2022). Last, the two candidate CTCF binding motifs that are within the forewing- enriched ATAC-seq peak are also conserved across Lepidoptera and Trichoptera (Fig. S1), two lineages that diverged around 300 Mya (Kawahara et al. 2019; Thomas et al. 2020).

CRISPR perturbation of Antp-Ubx_BE results in FW➞HW homeoses.

(A) Antp-Ubx_BE sgRNA targeting (cyan triangle) of a FW- enriched ATAC-peak (magenta) within the Ubx last intron. (B-C) Two examples of J. coenia Antp-Ubx_BE crispants showing mosaic FW➞HW homeoses, shown in dorsal views. CL-WT : contralateral, horizontally flipped images of forewings from the same individuals. WT HW : wild type hindwings from the same individual and mutant forewing side. Both individuals show disruption of their Radial veins (R1-R5 area). The specimen shown in C displays a partial, ectopic eyespot (asterisk). (D-E) Immunofluorescent detection of the UbdA epitope (green) in fifth instar wings disks of Antp-Ubx_BE crispants, revealing ectopic antigenicity in forewings. WT forewings of similar stage, and HW from the same crispant individuals, are shown for comparison as insets. Green autofluorescence was observed in tracheal tissues.

To test this hypothesis, we used CRISPR targeted mutagenesis to perturb Antp-Ubx_BE and assess its functionality, and designed a single sgRNA in a conserved sequence within the forewing- enriched ATAC-seq (Fig. S1). Remarkably, CRISPR mutagenesis of the Antp-Ubx_BE target induced G0 mutants with homeotic transformations of their forewings into hindwings (Figs. 4B-C and S2), including identity shifts in patterns, venation, and wing shape. It is important to note that none of the resulting crispants showed hindwing effects. Thus, we can reasonably attribute forewing homeotic phenotypes to indel mutations restricted to the intronic region, without disruption of the Ubx transcript, as this would result in hindwing phenotypes (Matsuoka and Monteiro 2021; Tendolkar et al. 2021).

Homeotic clones are first visible in Antp-Ubx_BE crispants at the pupal stage, with streaks of thinner cuticle, sometimes associated with local necrosis or with suture defects in the ventral midline, in particular where leg and wing pouches adjoin (Fig. S3). Color pattern homeotic clones were salient at the adult stage, with for example, clonal losses of the forewing specific white-band, and partial acquisitions of the large M1-M2 hindwing eyespot. In one specimen, an ectopic, partial M1-M2 hindwing eyespot appeared in the R5-M1 region, suggesting a perturbation of the eyespot induction process in this wing. Nymphalid forewings have five radial veins (R1-5), which provide sturdiness for flight (Wootton 1993), while hindwings have only two Radial veins. Forewing homeotic mutants showed mosaic venation defects in the Radial vein area (Fig. 4B). Finally, higher expressivity mutant forewings were smaller and rounder, reminiscent of hindwing shape.

Next, we dissected fifth instar larval wing disks from developing Antp-Ubx_BE crispants, and monitored the expression of Ubd-A (Ubx and Abd-A epitopes), normally restricted to the hindwing and only present in the forewing peripodial membrane (Weatherbee et al. 1999). Crispants showed forewing clones with strong ectopic expression of Ubd-A (Figs. 4D-E and S4). This result supports the inference that Antp-Ubx_BE forewing homeoses are due to the de-repression of Ubx in this tissue.

Mutational interrogation of lncRNA-encoding regions at the Ubx locus

We used CRISPR mutagenesis to test the function of the two lncRNA-encoding loci at the Ubx locus. Mutagenesis of the Ubx-IT1 first exon in J. coenia, and of the Ubx-T1 promoter in V. cardui, both resulted in crispants with small homeotic phenotypes in forewings and hindwings (Figs. 5 and S5). This result contrasts with Ubx exon mKO experiments, which only generate hindwing phenotypes (Tendolkar et al. 2021). Given the scarcity of Ubx-IT1 crispants obtained (11 out of 236 adults), and the small size of the homeotic clones within them, we infer the occasional phenotypes may be due to rare alleles. Thus, rather than evidence of functionality of the Ubx-IT1 transcript, the homeotic phenotypes may rather reflect the effects of regulatory perturbation on Ubx itself, with some random mutations in this intronic region resulting in hindwing Ubx loss-of-function, and some others triggering derepression in forewings. Likewise, next we mutagenized the first exon of Ubx-AS5’, located upstream of the Ubx promoter, and obtained twelve hindwing mutants and a single forewing mutant (Fig. 6 and S6). As with Ubx-IT1 CRISPR experiments, these results may be explained by regulatory disruption of Ubx transcription, with a higher ratio of hindwing phenotypes compared to forewings linked to the proximity of the Ubx promoter. Overall, we conclude that the mutational interrogation at these loci can result in dual loss (hindwing) and gain (forewing) of Ubx function effects. Deciphering whether or when these effects affected Ubx expression via local cis-regulatory modules, impairment of lncRNA transcripts, or larger indels overlapping with Ubx exons, will require further study (see Discussion).

Rare, dual homeoses obtained from CRISPR mutagenesis of the lncRNA_Ubx-IT1 5’ region.

(A) Genomic context of the sgRNA targets (here shown in J. coenia), in the promoter and first exon of the non-coding Ubx-IT1 transcript. (B-C) Dorsal and ventral views of a J. coenia crispant displaying dual homeoses, i.e. with both FW➞HW (presumably due to Ubx gain-of-expression), and HW➞FW clones (akin to Ubx null mutations). Insets on the right describe forewing mutant clones (IT1 mKO), in apposition to CL-WT (contralateral forewings from the same individual), and WT HW (wild type hindwings from the same individual and mutant forewing side). (D) Examples of dual homeoses obtained when targeting orthologous sites in V. cardui.

Homeoses obtained from CRISPR mutagenesis of the lncRNA Ubx-AS5’ first exon.

(A) CRISPR sgRNA targets (here shown in J. coenia), in the first exon of the non-coding Ubx-AS5’ transcript. (B) A single J. coenia crispant showed a FW➞HW transformation. Insets on the right describe forewing mutant clones (AS5’ mKO), in apposition to CL-WT (contralateral forewings from the same individual), and WT HW (wild-type hindwings from the same individual and mutant forewing side). (C-D) Examples of HW➞FW homeoses obtained in J. coenia or when targeting orthologous sites in V. cardui. Scale bars: 500 μm.

Dual effects of mutagenesis in a putative Ubx cis-regulatory module

In an attempt to probe for Ubx hindwing-specific regulatory sequences, we focused on a ∼ 25kb region in the first intron of Ubx that displays an ATAC-seq signature of hindwing enrichment in open- chromatin relative to forewings, hereafter dubbed CRM11 (Fig. 7A). We sub-divided this differentially accessible region into four peaks (11a, b, c and d). Targeting the ATAC-seq peaks with multiple sgRNAs spanning sub-domains 11a and 11c (UbxCRE11a2c5 in V. cardui, 11a2a3c5c6 in J. coenia), or with a single target in 11c (UbxCRE11c5 in V. cardui) yielded dual homeoses : FW➞HW and HW➞FW (Figs. 7B-D and S7). Hindwing effects were reminiscent of Ubx protein coding knockouts (Tendolkar et al. 2021), indicating that these crispant alleles with a hindwing phenotype produce Ubx loss-of-function effects.

CRISPR perturbation of Ubx CRM11 generates occasional dual homeotic phenotypes.

(A) Overview of ATAC-seq differential chromatin accessibility profiles (hindwing - head tissues, green ; forewing - head tissue, magenta) across the Ubx first exon. Several regions show differential opening between wings, one of which (CRM11), was targeted here for CRISPR perturbation (sites a2 and c5 indicate sgRNA targets). (B) Dual homeosis phenotypes obtained in V. cardui following dual-targeting of UbxCRE11a2c5, including homeoses in color patterns and scale morphology. (D) Additional example of a V. cardui UbxCRE11a2c5 crispant with a forewing phenotype (gain of hindwing hair patches, arrowheads). (E) Example of mild hindwing homeoses showing a white eyespot focus on the dorsal and ventral sides. These effects were previously shown to occur in coding Ubx CRISPR knock-out experiments (Tendolkar et al, 2021). Contralateral (CL) WT wings are shown for comparison with mutant wings (B-E). Colored dashed lines: wing veins. Scale bars: 500 μm.

Individuals with hindwing clones 2.75 times more common than individuals with forewings in this dataset. Similarly to the lncRNA loci perturbation experiments, dual homeoses may indicate the presence of hindwing activators and forewing repressors in the CRM11 region, with various CRISPR alleles producing a spectrum of indels and effects (see Discussion). It is noteworthy that while single- target experiments showed little lethality (55% hatching rate), dual or quadruple injection mixes resulted in low hatching rates of injected embryos (∼ 10%). Multiple targeting thus appears to induce high-rates of embryonic lethality, possibly due to chromosomal damage (Cullot et al. 2019; Zuccaro et al. 2020). Dual targeting with a2+c5 also yielded partial HW➞FW homeoses in V. cardui under the form of ectopic white eyespot foci phenotypes (Fig. 7E), as occasionally observed in Ubx null crispants (Tendolkar et al. 2021), seemingly due to hypomorphic or heterozygous allelic states.

Next, we focused on a single target shared between both V. cardui and J. coenia in the 11b sub- domain. A whole genome alignment between 23 lepidopteran species and 2 trichopteran species indicated that region 11b is deeply conserved, suggesting important functional constraints on its sequence (Fig. S8A-B). Mutagenesis of 11b yielded a relatively high hatching rate (mean = 51.8 %), indicating a rare occurrence of the deleterious mutational effects observed in multiple targeting (see above). Four J. coenia crispants and two V. cardui crispants were obtained, all exclusively showing hindwing phenotype devoid of forewing effects. HW➞FW homeoses included a variety of phenotypes all seen in Ubx CDS mutants (Tendolkar et al. 2021), including transformations of the orange Discalis elements and the white band in J. coenia, and partial shifts in eyespot identity (Fig. S8C). Together the consistency in direction of transformations and the deep conservation of the 11b region suggests it may encode an enhancer necessary for the transcriptional activation of Ubx in hindwings.

A sample of spontaneous homeotic mutants in Heliconius butterflies

Homeotic shifts between forewings and hindwings can occur naturally in Lepidoptera, and have been documented as pattern aberrations in museum specimens (Sibatani 1980, 1983). As a complement to CRISPR-induced homeoses, we document here a rich sample of forewing/hindwing homeotic mutants in the genus Heliconius, systematically collected by L. E. Gilbert between 1987 and 2016 in captive stocks at UT Austin, as well as in the wild. Across these 15 spontaneous mutants, 12 show HW➞FW clones (Fig. S9), against 3 specimens with FW➞HW effects (Fig.8). Mutant clones in this dataset were always posterior to the M2 vein. Only 2 mosaic phenotypes were found on a dorsal side, with the 13 others appearing ventrally. These homeotic mosaics show pattern shifts with complete fore/hindwing conversions of scale types, as seen for instance in the loss of gray wing coupling scales on posterior ventral forewings (Fig. 8A-B), or conversely, in their acquisition in posterior hindwings (arrowheads in Fig. S9B-D, H). Homeoses also include noticeable local changes in wing shape, particularly in hindwings (asterisks in Fig. S9). Taken together, these effects are akin to CRISPR-induced perturbations at the Ubx locus. We speculate that fore/hindwing homeotic aberrations, found in nature and captive stocks, result from mutations at the Ubx locus itself.

Mosaic forewing homeoses in Heliconius butterfly spontaneous mutants.

Wild-type and mutant sides from the same individuals are shown in each panel, with one side digitally flipped to match left-to-right orientation. A. Heliconius melpomene rosina, ventral view. Wild-caught in the Osa Peninsula (Costa Rica), October 1989. B. Heliconius cydno galanthus, ventral view (magnified inset in B’). Stock culture from Organisation for Tropical Studies station, La Selva (Costa Rica), June 1990 C. Heliconius himera, dorsal view (magnified inset in C’). Stock Culture in the butterfly farm Heliconius Butterfly Works in Mindo (Ecuador), March 2008.

Discussion

An intronic region with ATAC-seq hindwing-enrichment regulates Ubx

All CRISPR targets yielded homeotic phenotypes (Fig. 9), with two kinds of interference with Ubx expression – forewing gain-of-function effects, and hindwing loss-of-function effects – and indicating the presence of regulatory sequences (broadly defined), that repress or enhance Ubx expression in this region. It is crucial here to highlight the limitations of the method, in order to derive proper insights about the functionality of the regulatory regions we tested. In essence, butterfly CRISPR experiments generate random mutations by non-homologous end joining repair, that are usually deletions (Connahs et al. 2019; Mazo-Vargas et al. 2022; Van Belleghem et al. 2023), and they require genotyping in a second (G1) generation to be properly matched to a phenotype (genotyping G0 mosaic wings is limited, because adult wings lost scale building cells that underlie a given phenotype). Previous data from other organisms suggests that Cas9 nuclease targeting can generate larger than expected mutations that evade common genotyping techniques (Shin et al. 2017; Adikusuma et al. 2018; Kosicki et al. 2018; Cullot et al. 2019; Owens et al. 2019). Even under the assumption that such mutations are relatively rare in butterfly embryos, the fact we injected >100 embryos in each experiment makes their occurrence likely (Fig. 9).

Summary of wing homeosis phenotypes obtained from mutational interrogation.

(A) CRISPR targets at non-coding regions across the Ubx region, here visualized in J. coenia. (B) Summary of injection and adult phenotype data obtained across CRISPR experiments. FW/HW crispants : total number of individuals with forewing or hindwing homeotic clones, regardless of the injected species. Individuals with dual homeosis are counted in both categories. Nmut/Ninj : number of crispants obtained (Nmut), over the number of injected embryos for each species. Bold: experiments with consistent effects in only one segment. See Table 1 for details.

When targeting hindwing-enriched ATAC-seq peaks within the first intron of Ubx – from CRM11 to the hindwing-enriched open-chromatin peak that coincides with the first exon of Ubx-IT1 – we obtained a mixture of hindwing and forewing phenotypes. Given the potential heterogeneity of allele sizes in these experiments, it is difficult to conclude robustly about the function of individual targets. Nonetheless, the phenotypic data and in particular the obtention of dual homeoses suggest we disrupted sequences that are necessary to Ubx activation in hindwings, as well as to its repression in forewings. Bifunctional cis-regulatory elements that can switch between enhancer and silencer roles are prevalent in Drosophila (Gisselbrecht et al. 2020; Segert et al. 2021; Pang et al. 2022). The CRM11 and IT1 targets adjoin or overlap with open-chromatin signals in both wing sets (Figs. 5A and 7A), providing circumstantial evidence that these regions might serve as bifunctional elements. Similar observations were recently made in mutational interrogation experiments of the butterfly WntA patterning gene (Mazo-Vargas et al. 2022). Alternatively, silencers and enhancers may be tightly linked and interact in close proximity to shape gene expression (Méndez-González et al. 2023), implying in our case that forewing and hindwing phenotypes are mediated by alleles spanning adjacent but distinct elements. A formal test of these mechanisms would require germline transmission and genotyping of these alleles, which was unsuccessful in our attempts at crossing Ubx cis-regulatory crispants.

In contrast with the dual effects obtained when targeting CRM11a+c (Fig. 9), CRM11b perturbation resulted in hindwing-limited effects, and may suggest that an Ubx enhancer was consistently compromised in this specific dataset. The high lethality and small size of mutant wing streaks suggest that only individuals with sparse, small mutant mitotic clones can survive to the adult stage. If this is true, CRM11 may contain pleiotropic enhancers that are vital for normal Ubx function at earlier stages, but expression-reporter studies will be required to test this.

Parsing lncRNA-encoding regions – correlation or cause?

LncRNAs are emerging as important regulators of gene expression and developmental processes (Zhang et al. 2019; Statello et al. 2021). IT1 targeting generated a majority of forewing phenotypes, suggesting perturbation of Ubx repression in the T2 segment. However, IT1 showed low expression in forewing RNAseq datasets from K. inachus, and higher expression in the hindwing (Fig. 1B), a pattern inconsistent with a repressive role of the antisense IT1 transcript on Ubx expression. It is generally challenging to disentangle the effects of transcription of a non-coding element from the potential effects of adjacent enhancers (Natoli and Andrau 2012; Pease et al. 2013). Therefore, an alternative explanation would be that the phenotypes are confounded by the overlap and proximity to open- chromatin regions, which may play cis-regulatory roles on Ubx via DNA-protein interactions, rather than via the lncRNA. If this is the case, it is possible that the targeted Ubx-IT1 site, which yielded homeoses in both directions and bears both forewing and hindwing open-chromatin, is a bifunctional cis- regulatory element that can shift regulatory activities between these tissues (Gisselbrecht et al. 2020). Targeted mutagenesis of the Ubx-AS5’ first exon mainly generated hindwing phenotypes, although with a relatively low-efficiency. Because this target is about 10 kb away from the Ubx promoter itself, it is plausible that the observed phenotypes were due to large deletions reaching the promoter region of Ubx. Because mutational interrogation alone cannot discern if phenotypic effects arose from regulatory failure at the chromatin or transcript level, determining if AS5’ and IT1 are functional lncRNAs will need further examination.

A TAD boundary element likely acts as an insulator preventing Ubx forewing expression

Tight maintenance of TAD boundaries at the Hox locus is crucial for accurate segment identity and is facilitated by insulator proteins (Stadler et al. 2017; Gambetta and Furlong 2018; Ramírez et al. 2018). The Antp-Ubx_BE element we targeted is in a good position to block interactions between Antp and Ubx (Figs. 2-3). Consistent with this idea, the last intron of Ubx contains 8 CTCF binding motifs that are all clustered within 5-kb around the forewing-enriched ATAC peak, including two sites at highly conserved positions that are only 100-bp away from the CRISPR target (Fig. S1). CTCF sites prevent cross-talk between regulatory domains in the fly BX-C complex, and result in Hox misexpression when deleted (Postika et al. 2018; Kyrchanova et al. 2020; Kaushal et al. 2022; Kahn et al. 2023). Thus, the density of CTCF sites in this region may be indicative of a bona fide insulator active in forewings.

CRISPR mutagenesis of Antp-Ubx_BE generated FW➞HW homeoses associated with a gain of UbdA antigenicity in forewings, with no effects in the other direction, in stark contrast with other targets (Fig. 9B). This suggests a possible de-insulation of the TAD boundary in the crispant clones, resulting in a TAD fusion or in a long-range interaction between a T2-specific enhancer and Ubx promoter. Similar de-insulating effects of deletion alleles have been described at the Notch locus in Drosophila (Arzate- Mejía et al. 2020), in digit-patterning mutants in mice and humans (Lupiáñez et al. 2015; Anania et al. 2022), or at murine and fly Hox loci depleted of CTCF-mediated regulatory blocking (Narendra et al.

2015; Gambetta and Furlong 2018; Kyrchanova et al. 2020). It will be interesting to profile the binding of insulator proteins and transcriptional repressors across the butterfly Hox TAD landscape to shed more light onto the mechanisms of Ubx insulation, using in vivo assays (Bowman et al. 2014), or in silico predictions that take advantage of updated binding matrices for insect insulator proteins (Mitra et al. 2018). Of note, our CRISPR target is adjacent to an hindwing-enriched peak that also presented CTCF binding sites (Fig. 4A). Following a similar logic, this site could be a candidate insulator specific to Ubx-expressing tissues like the hindwing, a hypothesis that will require further testing.

Making sense of spontaneous wing homeotic mutants

In this article, we documented a large sample of spontaneous homeotic mutants obtained in Heliconius spp. All homeotic clones were limited to the wing posterior compartments (ie. posterior to the M2 vein), possibly because of parasegmental, compartment-specific regulatory domains that played historic roles in the study of Drosophila BX-C regulation (Maeda and Karch 2015). Sibatani documented in Lepidoptera that “the mosaics of F/H homeosis tend to occur most frequently in the posterior half of the wing, the boundary of the anterior and posterior halves occurring somewhere in space M1-M2” (Sibatani 1983). Our collection of spontaneous Heliconius mutants only displayed clones in posterior regions, consistently with this trend. However, our CRISPR perturbation assays of J. coenia and V. cardui cis-regulatory regions also generated anterior clones, with all targets. Deciphering how butterfly Ubx regulation is compartmentized between parasegmental or wing antero-posterior domains will require additional investigation. Most Heliconius homeoses were in the hindwings (ie. putative Ubx loss-of- expression clones), and among these, all but one were ventral (Fig. S9). Three mutants showed forewing homeoses (ie. putative Ubx gain-of-expression clones), two of them ventral and one of them dorsal (Fig. 8). The systematic reviews of wing homeosis in Lepidoptera found that forewing homeoses are almost as common as hindwing ones (Sibatani 1980, 1983). Our mutational interrogation assays, while coarse in nature, revealed the existence of activating and repressing cis-regulatory sequences at the Ubx locus itself. Spontaneous FWHW homeoses observed in butterflies and moths may thus result from somatic mutations or transposition events at this locus.

Materials and methods

Genome annotations and transcriptomic analysis

Nymphalid genome sequences of the Hox cluster and their annotations were extracted from the NCBI Assembly and Lepbase online repositories (Challis et al. 2016; Kitts et al. 2016) as follows : V. cardui from NCBI ilVanCard2.1 and LepBase Vc_v1 ; A. (Nymphalis) io from NCBI ilAglIoxx1.1; J. coenia from Lepbase Jc_v2; P xylostella from NCBI ilPluXylo3.1. The Ubx regions from ilVanCard2.2, Vc_v1, and Jc_v2 were manually re-annotated using wing transcriptome data on the NCBI SRA (BioProjects PRJNA661999, PRJNA293289, PRJNA237755, PRJNA385867, and PRJNA498283) The genome sequence of K. inachus was obtained from the Dryad repository (Yang et al. 2020). Differential gene expression analysis across the K. inachus Ubx locus were carried out using wing transcriptome data available on the NCBI SRA (BioProject PRJNA698433), following a manual re-annotation of a preliminary gene models provided by Peiwen Yang and Wei Zhang (Wang et al. 2022). All transcripts analyses were performed using the STAR intron-aware aligner and DEseq2 expression analysis package as previously described (Love et al. 2014; Dobin and Gingeras 2016; Hanly et al. 2019, 2022). Expression levels were calculated as genome-wide normalized counts and pairwise Wald tests were performed to assess differential expression between forewings and hindwings. Multiple test adjustment was performed using Benjamini and Hochberg correction.

Hi-C and ATAC-seq analyses

Hi-C data from J. coenia fifth instar larval forewings and 48-72 h APF pupa hindwings are available at the NCBI SRA BioProject PRJNA641138 (van der Burg et al. 2020). Triplicated ATAC-seq datasets for larval and pupal wing and head tissues of J. coenia and V. cardui (van der Burg et al. 2019, 2020; Mazo- Vargas et al. 2022) are available on the NCBI SRA BioProjects PRJNA497878, PRJNA695303, and PRJNA559165. All the ATAC-seq and Hi-C data were re-analysed on J. coenia and V. cardui Ubx genomic regions as previously described (Mazo-Vargas et al. 2022). Briefly, matrices of interactions were constructed by mapping paired reads against the Junonia coenia genome (Mazo-Vargas et al., 2022) using hicBuildMatrix (Ramírez et al. 2018). Matrices from larvae and pupae were normalized using hicNormalize and corrected with the Knight-Ruiz matrix balancing algorithm.The definitions of topologically associating domains (TADs) can be influenced by various factors such as the choice of software, parameters, sequencing depth, and the presence of experimental noise. To ensure reliability, it is recommended to compare TAD calls with independent datasets, such as histone marks or known factors associated with TAD boundaries. In the absence of these specific datasets, we employed a different combination of parameters in the hicFindTADs tool and compared the resulting TAD calls. HiC matrices at 10 kb and 20 kb bin resolutions were utilized, and TAD insulation scores were evaluated using a false-discovery rate correction for multiple testing, with p-value thresholds of 0.01 and 0.005. Consistent TAD boundary calls with negative TAD separation scores were selected to define domain limits at 10 kb and 20 kb matrix resolutions.

CTCF motif binding predictions

The program fimo was used to scan for the J. coenia candidate TAD boundary region (HiC_scaffold_12:6430000-6444000) for canonical CTCF binding sites, using the positional weight matrix MA0205.1 deposited in the JASPAR database (Holohan et al. 2007; Cuellar-Partida et al. 2012; Castro-Mondragon et al. 2022).

Genomic conservation analyses

We generated whole-genome alignments of 25 Lepidoptera and 2 Trichoptera reference species from NCBI Assembly using ProgressiveCactus (Armstrong et al. 2020), and HAL tools (Hickey et al. 2013) for converting the resulting HAL file to the MAF format. We provided a species topology tree of 23 Lepidoptera species to PhyloFit (Hubisz et al. 2011) to fit a multiple sequence alignment on the reference J. coenia Ubx locus, using HKY85 as the substitution model. Using PhastCons (Siepel et al. 2005), we then generated conservation score plots using standard parameters (target-coverage = 0.45; expected-length = 12; rho = 0.1) stored in BED and WIG file formats.

Butterfly rearing and CRISPR microinjections

J. coenia and V. cardui colonies were maintained at 25°C and 60-70% relative humidity in a growth chamber with a 14:10 light:dark photoperiod. Larval rearing on artificial diets, egg collection, and microinjections followed previously described methods (Martin et al. 2020; Tendolkar et al. 2021). Cas9:sgRNA heteroduplexes were prepared as previously described (Martin et al., 2020). Frozen aliquots of Cas9-2xNLS (2.5 μL ; 1,000 ng/μL) and sgRNA (2.5 μL ; 500 ng/μL) were mixed in 2:1 and 4:1:1 mass ratios for single and dual target injections, respectively. CRISPR sgRNA targets are listed in Table S1.

Antibody stainings

Fifth instar wing disks were dissected in ice cold Phosphate Buffer Saline (PBS), fixed for 15-20 min at room temperature in methanol-free formaldehyde diluted to 4% in PBS / 2mM EGTA (egtazic acid), washed in PBS with 0.1% Triton X-100 (PT), stored in PT with 0.5% Bovine Serum albumin (PT-BSA), incubated overnight at 4°C in PT-BSA with a 1:5 dilution of the FP6.87 antibody serum (mouse monoclonal, Developmental Studies Hybridoma Bank), and washed in PT. A 1:250 dilution of anti-Mouse IgG antibody coupled to AlexaFluor488 or Rabbit AlexaFluor555 was made in PT-BSA and spun down at 14,000 rcf to remove aggregates, and incubated with wings for 2 h at room temperature, before additional washes, incubation in 50% glycerol-PBS with DAPI (4′,6-diamidino-2-phenylindole) nuclear stain, and incubation and mounting in 60% glycerol-PBS with 2mM of EDTA (Ethylenediaminetetraacetic acid).

Imaging

Full-mount photographs of J. coenia and V. cardui were taken on a Nikon D5300 digital camera mounted with an AF-S VR MicroNikkor 105mm f/2.8G lens, with magnified views taken using a Keyence VHX-5000 digital microscope mounted with VH-Z00T and VH-Z100T lenses. Immunofluorescent stainings were imaged on an Olympus BX53 epifluorescent microscope mounted with UPLFLN 4x/0.13 and 10X/0.3 objectives.

eLife Assessment

This valuable paper examines the Bithorax complex in several butterfly species, in which the complex is contiguous and not split, as it is in the well-studied fruit fly Drosophila. Based on genetic screens and genetic manipulations of a boundary element involved in segment-specific regulation of Ubx, the authors provide solid evidence for their conclusions, which could be further strengthened by additional data and analyses. The data presented are relevant for those interested in the evolution and function of Hox genes and of gene regulation in general.

Reviewer #1 (Public Review):

In their article, "Cis-regulatory modes of Ultrabithorax inactivation in butterfly forewings," Tendolkar and colleagues explore Ubx regulation in butterflies. The authors investigated how Ubx expression is restricted to the hindwing in butterflies through a series of genomic analyses and genetic perturbations. The authors provide evidence that a Topoologiacally Associated Domain (TAD) maintains a hindwing-enriched profile of chromatin around Ubx, largely through an apparent boundary element. CRISPR mutations of this boundary element led to ectopic Ubx expression in forewings, resulting in homeotic transformation in the wings. The authors also explore the results of the mutation in two non-coding RNA regions as well as a possible enhancer module. Each of these induces homeotic phenotypes. Finally, the authors describe a number of homeotic phenotypes in butterflies, which they relate to their work.

Together, this was an interesting paper with compelling initial data. That said, I have several items that I feel would warrant further discussion, presentation, or data.

First, I would not state, "Little is known about how Hox genes are regulated outside of flies." They should add "in insects" since so much in known in vertebrates

For Figure 1, it would aid the readers if the authors could show the number of RNAseq reads across the locus. This would allow the readership to evaluate the frequency of the lncRNAs, splice variants, etc.

How common are boundary elements within introns? Typically, boundary elements are outside gene bodies, so this could be explored further. This seems like an interesting bit of biology which, following from the above point, it would be interesting to, at a minimum, discuss, but also relate to how transcription occurs through a possible boundary element (are there splice variants, for example?).

The CRISPR experiments led to compelling phenotypes. However, as a Drosophila biologist, I found it hard to interpret the data from mosaic experiments. For example, in control experiments, how often do butterflies die? Are there offsite effects? It's striking that single-guide RNAs led to such strong effects. Is this common outside of this system? Is it possible to explore the function effects at the boundary element - are these generating large deletions (for example, like Mazo-Vargas et al., 2022)?

For the mosaic experiments, how frequent are these effects in nature or captive stocks? Would it be possible to resequence these types of effects? At the moment, this data, while compelling, was hard to put into the context of the experiments above without understanding how common the effects are. Ideally, there would be resequencing of these tissues, which could be targeted, but it was not clear to me the general rates of these variants.

In sum, I enjoyed the extensive mosaic perturbations. However, I feel that more molecular descriptions would elevate the work and make a larger impact on the field.

Reviewer #2 (Public Review):

Summary:

The existence of hox gene complexes conserved in animals with bilateral symmetry and in which the genes are arranged along the chromosome in the same order as the structures they specify along the anteroposterior axis of organisms is one of the most spectacular discoveries of recent developmental biology. In brief, homeotic mutations lead to the transformation of a given body segment of the fly into a copy of the next adjacent segment. For the sake of understanding the main observation of this work, it is important to know that in loss-of-function (LOF) alleles, a given segment develops like a copy of the segment immediately anterior to it, and in gain-of-function mutations (GOF), the affected segment develops like a copy of the immediately posterior segment. Over the last 30 years the molecular lesions associated with GOF alleles led to a model where the sequential activation of the hox genes along the chromosome result from the sequential opening of chromosomal domains. Most of these GOF alleles turned out to be deletions of boundary elements (BE) that define the extent of the segment-specific regulatory domains. The fruit fly Drosophila is a highly specialized insect with a very rapid mode of segmentation. Furthermore, the hox clusters in this lineage have split. Given these specificities it is legitimate to question whether the regulatory landscape of the BX-C we know of in D.melanogaster is the result of very high specialization in this lineage, or whether it reflects a more ancestral organization. In this article, the authors address this question by analyzing the continuous hox cluster in butterflies. They focus on the intergenic region between the Antennapedia and the Ubx gene, where the split occurred in D.melanogaster. Hi-C and ATAC-seq data suggest the existence of a boundary element between 2 Topologically-Associated-Domain (TAD) which is also characterized by the presence of CTCF binding sites. Butterflies have 2 pairs of wings originating from T2 (forewing) specified by Antp and T3 specified by Ubx (hindwing). Remarkably, CRISPR mutational perturbation of this boundary leads to the hatching of butterflies with homeotic clones of cells with hindwings identities in the forewing (a posteriorly oriented homeotic transformation). In agreement with this phenotype, the authors observe ectopic expression of Ubx in these clones of cells. In other words, CRISPR mutagenesis of this BE region identified by molecular tool give rise to homeotic transformations directed towards more posterior segment as the boundary mutations that had been 1st identified on the basis of their posterior oriented homeotic transformation in Drosophila. None of the mutant clones they observed affect the hindwing, indicating that their scheme did not affect the nearby Ubx transcription unit. This is reassuring and important first evidence that some of the regulatory paradigms that have been proposed in fruit flies are also at work in the common ancestor to Drosophilae and Lepideptora.

Given the large size of the Ubx transcription unit and its associated regulatory regions it is not surprising that the authors have identified ncRNA that are conserved in 4 species of Nymphalinae butterflies, some of which also present in D.melanogaster. Attempts to target the promoters by CRISPR give rise to clones of cells in both forewings and hindwings, suggesting the generation of regulatory mutations associated with both LOF and GOF transformations. The presence of clones with dual homeosis suggests the targeting of Ubx activator and repression CRMs. Unfortunately, these experiments do not allow us to make further conclusions on the role of these ncRNA or in the identification of specific regulatory elements. To the opinion of this reviewer, some recent papers addressing the role that these ncRNA may play in boundary function should be taken with caution, and evidence that ncRNA(s) regulate boundaries in the BX-C in a WT context is still lacking.

Strengths:

The convincing GOF phenotype resulting from the targeting of the Antp-Ubx_BE.

Weaknesses:

The lack of comparisons with the equivalent phenotypes obtained in D.melanogaster with for example the Fub mutation.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

This valuable paper examines the Bithorax complex in several butterfly species, in which the complex is contiguous and not split, as it is in the well-studied fruit fly Drosophila. Based on genetic screens and genetic manipulations of a boundary element involved in segment-specific regulation of Ubx, the authors provide solid evidence for their conclusions, which could be further strengthened by additional data and analyses. The data presented are relevant for those interested in the evolution and function of Hox genes and of gene regulation in general.

We are deeply grateful to the eLife editorial team and the two reviewers for their thoughtful and constructive feedback. We have used this feedback to improve our manuscript and have provided a point-by-point response below.

Public Reviews:

Reviewer #1 (Public Review):

In their article, "Cis-regulatory modes of Ultrabithorax inactivation in butterfly forewings," Tendolkar and colleagues explore Ubx regulation in butterflies. The authors investigated how Ubx expression is restricted to the hindwing in butterflies through a series of genomic analyses and genetic perturbations. The authors provide evidence that a Topologically Associated Domain (TAD) maintains a hindwing-enriched profile of chromatin around Ubx, largely through an apparent boundary element. CRISPR mutations of this boundary element led to ectopic Ubx expression in forewings, resulting in homeotic transformation in the wings. The authors also explore the results of the mutation in two non-coding RNA regions as well as a possible enhancer module. Each of these induces homeotic phenotypes. Finally, the authors describe a number of homeotic phenotypes in butterflies, which they relate to their work.

Together, this was an interesting paper with compelling initial data. That said, I have several items that I feel would warrant further discussion, presentation, or data.

First, I would not state, "Little is known about how Hox genes are regulated outside of flies." They should add "in insects" since so much in known in vertebrates

Corrected

For Figure 1, it would aid the readers if the authors could show the number of RNAseq reads across the locus. This would allow the readership to evaluate the frequency of the lncRNAs, splice variants, etc.

We have found it useful in the past to feature “Sashimi Plots”, as they provide a good overview of transcript splicing junctions and read support. Here we could not accommodate this in our Fig. 1A as this would require compiling the RNAseq reads from many tissues and stages to be meaningful, and we would lose the resolution on forewing vs hindwing tissues that is important in this article (only the Kallima inachus dataset allows this comparison, and was used in Fig 1B). More specifically, the wing transcriptomes available for J. coenia and V. cardui are not deep enough to provide a good visualization of Antp alternative promoter usage or on AS5’ transcription.

How common are boundary elements within introns? Typically, boundary elements are outside gene bodies, so this could be explored further. This seems like an interesting bit of biology which, following from the above point, it would be interesting to, at a minimum, discuss, but also relate to how transcription occurs through a possible boundary element (are there splice variants, for example?).

We do not see evidence of alternative splicing, and prefer to avoid speculating on transcriptional effects, but we agree that the intragenicity of the TAD boundary is interesting. We briefly highlighted this point in the revised Discussion:

"Lastly, it is worth noting that the Antp/Ubx TAD boundary we identified is intragenic, within the last intron of Ubx. It is unclear if this feature affects Ubx transcription, but this configuration might be analogue to the Notch locus in Drosophila, which includes a functional TAD boundary in an intronic position (Arzate-Mejía et al. 2020)."

The CRISPR experiments led to compelling phenotypes. However, as a Drosophila biologist, I found it hard to interpret the data from mosaic experiments. For example, in control experiments, how often do butterflies die? Are there offsite effects? It's striking that single-guide RNAs led to such strong effects. Is this common outside of this system? Is it possible to explore the function effects at the boundary element - are these generating large deletions (for example, like Mazo-Vargas et al., 2022)? For the mosaic experiments, how frequent are these effects in nature or captive stocks? Would it be possible to resequence these types of effects? At the moment, this data, while compelling, was hard to put into the context of the experiments above without understanding how common the effects are. Ideally, there would be resequencing of these tissues, which could be targeted, but it was not clear to me the general rates of these variants.

We agree with this assessment completely: mosaics complicate the proper interpretation of CRISPR based perturbation assays in regulatory regions. Here, unlike in Mazo-Vargas et al. (2022), we were unable to breed homeotic effects to a G1 generation, possibly because the phenotypes are dominant and lethal at the embryonic stage (see also our reply to Reviewer 2). This means that mosaic mutants are often survivors with clones of restricted size in the wing, and they are probably rare, but we are unable to meaningfully measure a mutation spectrum frequency (e.g. how often large deletions are generated). As mentioned in the first paragraph of our Discussion, we think that many of the phenotypes we observed (besides the Ubx GOF effects from the BE targeting) were confounded by alleles that could include large SVs. We aim to address these questions in an upcoming manuscript, at a locus where regulatory perturbation does not impact survival, including using germline mutants and unbiased genotyping (whole genome resequencing).

We elaborated on this issue in our Discussion:

"It is crucial here to highlight the limitations of the method, in order to derive proper insights about the functionality of the regulatory regions we tested. In essence, butterfly CRISPR experiments generate random mutations by non-homologous end joining repair, that are usually deletions (Connahs et al. 2019; Mazo-Vargas et al. 2022; Van Belleghem et al. 2023). Ideally, regulatory CRISPR-induced alleles require genotyping in a second (G1) generation to be properly matched to a phenotype (Mazo-Vargas et al. 2022). Possibly because of lethal effects, we failed to pass G0 mutations to a G1 generation for genotyping, and were thus limited here to mosaic analysis. As adult wings have lost scale building cells that may underlie a given phenotype, we circumvented this issue by genotyping a pupal forewing displaying an homeotic phenotype in the more efficient Antp-Ubx_BE perturbation experiment (Fig. S4). In this case, PCR amplification of a 600 bp fragment followed by Sanger sequencing recovered signatures of indel variants, with mixed chromatograms starting at the targeted sites. But in all other experiments (CRM11, IT1, and AS5’ targets), we did not genotype mutant tissues, as they were only detected in adult stages and generally with small clone sizes. Some of these clones may have been the results of large structural variants, as data from other organisms suggests that Cas9 nuclease targeting can generate larger than expected mutations that evade common genotyping techniques (Shin et al. 2017; Adikusuma et al. 2018; Kosicki et al. 2018; Cullot et al. 2019; Owens et al. 2019). Even under the assumption that such mutations are relatively rare in butterfly embryos, the fact we injected >100 embryos in each experiment makes their occurrence likely (Fig. 9), and we are unable to assign a specific genotype to the homeotic effects we obtained in CRM11, IT1 and AS5’ perturbation assays."

Our revision also includes a new Fig. S4 that features the mosaic genotyping of a G0 Antp-Ubx_BE mutant tissue. While this does not fully address the reviewer questions, it provides reasonable validation that the frequent GOF effects we observed upon perturbation at this target site are generated by on-target indels from DNA repair.

Validation of CRISPR-induced DNA Lesions in an Antp-Ubx_BE crispant pupat forewing.

(A-A') Pupal forewing cuticle phenotype of an Antp-Ubx_BE J. coenia crispant, as in Fig. S3. (B-B") Aspect of the same forewing under trans-illumination following dissection out of the pupal case. Regions from mutant clones have a more transparent appearance. (C). Sanger sequencing of an amplicon targeting the Antp-Ubx_BE region in the mutant tissue shown in panel B", compared to a control wing tissue, showing mixed chromatogram around the expected CRISPR cutting site due to indel mutations from non-homologous end-joining.

In sum, I enjoyed the extensive mosaic perturbations. However, I feel that more molecular descriptions would elevate the work and make a larger impact on the field.

Reviewer #2 (Public Review):

Summary:

The existence of hox gene complexes conserved in animals with bilateral symmetry and in which the genes are arranged along the chromosome in the same order as the structures they specify along the anteroposterior axis of organisms is one of the most spectacular discoveries of recent developmental biology. In brief, homeotic mutations lead to the transformation of a given body segment of the fly into a copy of the next adjacent segment. For the sake of understanding the main observation of this work, it is important to know that in loss-of-function (LOF) alleles, a given segment develops like a copy of the segment immediately anterior to it, and in gain-of-function mutations (GOF), the affected segment develops like a copy of the immediately posterior segment. Over the last 30 years the molecular lesions associated with GOF alleles led to a model where the sequential activation of the hox genes along the chromosome result from the sequential opening of chromosomal domains. Most of these GOF alleles turned out to be deletions of boundary elements (BE) that define the extent of the segment-specific regulatory domains. The fruit fly Drosophila is a highly specialized insect with a very rapid mode of segmentation. Furthermore, the hox clusters in this lineage have split. Given these specificities it is legitimate to question whether the regulatory landscape of the BX-C we know of in D.melanogaster is the result of very high specialization in this lineage, or whether it reflects a more ancestral organization. In this article, the authors address this question by analyzing the continuous hox cluster in butterflies. They focus on the intergenic region between the Antennapedia and the Ubx gene, where the split occurred in D.melanogaster. Hi-C and ATAC-seq data suggest the existence of a boundary element between 2 Topologically-Associated-Domain (TAD) which is also characterized by the presence of CTCF binding sites. Butterflies have 2 pairs of wings originating from T2 (forewing) specified by Antp and T3 specified by Ubx (hindwing). Remarkably, CRISPR mutational perturbation of this boundary leads to the hatching of butterflies with homeotic clones of cells with hindwings identities in the forewing (a posteriorly oriented homeotic transformation). In agreement with this phenotype, the authors observe ectopic expression of Ubx in these clones of cells. In other words, CRISPR mutagenesis of this BE region identified by molecular tool give rise to homeotic transformations directed towards more posterior segment as the boundary mutations that had been 1st identified on the basis of their posterior oriented homeotic transformation in Drosophila. None of the mutant clones they observed affect the hindwing, indicating that their scheme did not affect the nearby Ubx transcription unit. This is reassuring and important first evidence that some of the regulatory paradigms that have been proposed in fruit flies are also at work in the common ancestor to Drosophilae and Lepidoptera.

Given the large size of the Ubx transcription unit and its associated regulatory regions it is not surprising that the authors have identified ncRNA that are conserved in 4 species of Nymphalinae butterflies, some of which also present in D.melanogaster. Attempts to target the promoters by CRISPR give rise to clones of cells in both forewings and hindwings, suggesting the generation of regulatory mutations associated with both LOF and GOF transformations. The presence of clones with dual homeosis suggests the targeting of Ubx activator and repression CRMs. Unfortunately, these experiments do not allow us to make further conclusions on the role of these ncRNA or in the identification of specific regulatory elements. To the opinion of this reviewer, some recent papers addressing the role that these ncRNA may play in boundary function should be taken with caution, and evidence that ncRNA(s) regulate boundaries in the BX-C in a WT context is still lacking.

Strengths:

The convincing GOF phenotype resulting from the targeting of the Antp-Ubx_BE.

Weaknesses:

The lack of comparisons with the equivalent phenotypes obtained in D.melanogaster with for example the Fub mutation.

We are grateful for this excellent contextualization of our findings and have incorporated some of the historical elements into our revision, as detailed below.

Reviewer #2 (Recommendations For The Authors):

In the whole paper, the authors bring the notion of boundaries through the angle of the existence of TADs and ignore almost entirely to explain the characteristics of boundary mutation in the BX-C. To my knowledge examples where targeted boundary deletions between TADs result in misregulation of the neighboring genes, and/or a phenotype, are extremely sparse (especially in the context of the mouse hox genes). Given the extensive litterature describing the boundary mutations and their associated GOF phenotypes, the paper would certainly gain strength if the authors justify their approach through this wealth of information. I must admit that this referee is surprised by the absence of any references to the founding work of the Karch and Bender laboratories on this topic. As a matter of fact, one of the founding members of the boundary class of regulatory elements was already brought in 1993 with the Fab-7 and Mcp elements of the BX-C. Based on gain-of-function homeotic phenotypes, additional Fab boundaries were added to the list. Finally, in 2013, Bender and Lucas (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3606092/) identified the Fub boundary element that delimits the Ubx and abd-A domains in the BX-C. Fub fulfills the criterium of lying at the border of 2 neighboring TADs. Significantly, a deletion of Fub leads to a very penetrant and strong homeotic gain-of-function phenotype in which the flies hatch with a 1st abdominal segment transformed into the 2nd. In agreement with this, abd-A is expressed one parasegment too anterior in embryos. This is exactly the observation gathered from the targeted mutations in the Antp-Ubx_BE; a dominant transformation of anterior to posterior wing accompanied by an ectopic expression of Ubx in the forming primordia of the forwing where it is normally silenced. I believe the paper would gain credibility if the results were reported with the knowledge of the similarities with Fub.

Line 53, I am not aware of the existence of TADs for each of the 9 regulatory domains. The insulators delimit the extent of the regulatory domains but certainly not of TADs.

We thank the reviewer for these suggestions, as well as for the correction – we agree our previous text suggested that all BX-C boundaries are TAD boundaries, which was incorrect. We added a new introduction paragraph that combines classic literature on GOF mutations at boundary elements with recent evidence these are TAD insulators, including Fub (as suggested), and adding Fab-7 for breadth of scope.

"For instance, the deletion of a small region situated between Ubx and abd-A produces the Front-ultraabdominal phenotype (Fub) where the first abdominal segment (A1) is transformed into a copy of the second abdominal segment A2, due to a gain-of-expression of abd-A in A1 where it is normally repressed (Bender and Lucas 2013). At the molecular level, the Fub boundary is enforced by insulating factors that separate Topologically Associating Domains (TADs) of open-chromatin, while also allowing interactions of Ubx and abd-A enhancers with their target promoters (Postika et al. 2018; Srinivasan and Mishra 2020). Likewise, the Fab-7 deletion, which removes a TAD boundary insulating abd-A and Abd–B (Moniot-Perron et al. 2023), transforms parasegment 11 into parasegment 12 due to an anterior gain-of-expression of Abd-B (Gyurkovics et al. 1990). By extrapolation, one may expect that if the Drosophila Hox locus was not dislocated into two complexes, Antp and Ubx 3D contact domains would be separated by a Boundary Element (BE), and that deletions similar with Fub and Fab-7 mutations would result in gain-of-function mutations of Ubx that could effectively transform T2 regions into T3 identities."

A reference to the 1978 Nature article of Lewis should be added after line 42 of introduction.

Added

Line 56-57; the BX-C encoded miRNAs are known to regulate Ubx and abd-A, but not Abd-B.

Corrected

From lines 57 to 61, the authors mention reports aimed at demonstrating a role of ncRNA into Ubx regulation. To my eyes, these gathered evidences are rather weak. A reference to the work of Pease et al in Genetics in 2013 should be mentioned (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3832271/).

Added. Our paragraph includes qualifier language about the functionality of the Ubx-related ncRNAs (“are thought to”, “appears to”), and updated references regarding bxd (Petruk et al. 2006; Ibragimov et al. 2023).

Line 62 authors, should write "Little is known about how Hox genes are regulated outside of Drosophila" and not flies.

Corrected

Lines 110-112 could lncRNA:Ubx-IT1 correspond to PS4 antisense reported by Pease et al in 2023 (see URL above)? Lines 115-117, could lncRNA:UbxAS5' correspond to bxd antisense of Pease et al in 2023 (see above)?

As we could not detect sequence similarities, we preferred to avoid drawing homology, and we intentionally avoided reference to the fly transcripts when we named IT1 and AS5’. This said, we agree it is important to clarify that further studies are needed to clarify this relationship. We elaborated on this point in our discussion:

"Of note, a systematic in-situ survey (Pease et al. 2013) showed that Drosophila embryos express an antisense transcripts in its 5’ region (lncRNA:bxd), as well as within its first intron (lncRNA:PS4). It is thought that Drosophila bxd regulates Ubx, possibly by transcriptional interference or by facilitation of the Fub-1 boundary effect (Petruk et al. 2006; Ibragimov et al. 2023), while the possible regulatory roles of PS4 remain debated (Hermann et al. 2022). While these dipteran non-coding transcripts lack detectable sequence similarity with the lepidopteran IT1 and AS5’ transcripts, further comparative genomics analyses of the Ubx region across the holometabolan insect phylogeny should clarify the extent to which Hox cluster lncRNAs have been conserved or independently evolved."

Lines 154-155: "This concordance between Hi-C profiling and CTCF motif prediction thus indicates that Antp-Ubx_BE region functions as an insulator between regulatory domains of Antp and Ubx ». This is only correlative, I would write "suggests" instead of "indicates" and add a "might function".

Corrected as suggested.

Line 254, I assume the authors wish to write Ubx-IT1 in V. cardui instead of Ubx-T1.

Typo corrected

Line 255 : Fig.5 is absent from the pdf file and replaced by table 1. I did not find a legend for Table 1.

Corrected, with our sincere apologies for the loss of this image in our first submission.

Line 293 "Individual with hindwing clones 2.75 times more common than...." "are" is missing?

Corrected

Lines 303-313, it is not entirely clear how many guide RNAs were injected. Would be useful to indicate the sites targeted in Fig.S8.

We specify in the revised text : using a single guide RNA (Ubx11b9)

Lines 323-337: it is not entirely clear to this referee (a drosophilist) if those spontaneous mutations can be inbred or whether these individuals are occasional mosaics. In general, did anyone try to derive lines from those mosaic animals? Is it possible to hit the germline at the syncitial stages at which the guides are injected? Are the individuals with wing phenotype fertile? Given the fact that the Antp-Ubx_BE mutations should be dominant, I wonder if this characteristic would not help in identifying germline transmission. Similar remark for the discussion where the authors explain at line 360, that genotyping can only be done in the progeny of the Go. I do not have the impression that the authors have performed this genotyping and if I am right, I do not understand why.

We improved our discussion section on this topic (new text in orange):

"It is crucial here to highlight the limitations of the method, in order to derive proper insights about the functionality of the regulatory regions we tested. In essence, butterfly CRISPR experiments generate random mutations by non-homologous end joining repair, that are usually deletions (Connahs et al. 2019; Mazo-Vargas et al. 2022; Van Belleghem et al. 2023). Ideally, regulatory CRISPR-induced alleles require genotyping in a second (G1) generation to be properly matched to a phenotype (Mazo-Vargas et al. 2022). Possibly because of lethal effects, we failed to pass G0 mutations to a G1 generation for genotyping, and were thus limited here to mosaic analysis. As adult wings have lost scale building cells that may underlie a given phenotype, we circumvented this issue by genotyping a pupal forewing displaying an homeotic phenotype in the more efficient Antp-Ubx_BE perturbation experiment (Fig. S4). In this case, PCR amplification of a 600 bp fragment followed by Sanger sequencing recovered signatures of indel variants, with mixed chromatograms starting at the targeted sites. But in all other experiments (CRM11, IT1, and AS5’ targets), we did not genotype mutant tissues, as they were only detected in adult stages and generally with small clone sizes. Some of these clones may have been the results of large structural variants, as data from other organisms suggests that Cas9 nuclease targeting can generate larger than expected mutations that evade common genotyping techniques (Shin et al. 2017; Adikusuma et al. 2018; Kosicki et al. 2018; Cullot et al. 2019; Owens et al. 2019). Even under the assumption that such mutations are relatively rare in butterfly embryos, the fact we injected >100 embryos in each experiment makes their occurrence likely (Fig. 9), and we are unable to assign a specific genotype to the homeotic effects we obtained in CRM11, IT1 and AS5’ perturbation assays."

We agree that the work we conducted with mosaics has important caveats. So far, our attempts at breeding homeotic G0 mutants have not been fruitful at this locus, while less deleterious loci can yield viable alleles into further generations, such as WntA (published) and cortex (in prep.). We prefer to stay vague about negative data here, as it is difficult to disentangle if they were due to real mutational effects (e.g. the alleles can be dominant and lethal in the G1 generation) to failure to germline carriers of mutations as founders, or to health issues that are often amplified by inbreeding depression (including a possible iflavirus in our V. cardui cultures).

We concur with the prediction that Antp-Ubx_BE mutations are probably dominant, and intend to follow up with similar GOF experiments in the Plodia pantry moth, a laboratory model for lepidopteran functional genomics that is more amenable than butterflies to inbreeding and long-term studies in mutant lines. In our experience (https://www.frontiersin.org/articles/10.3389/fevo.2021.643661/full), Ubx coding knock-out can be more extensive in Plodia than in butterflies, so we think these animals will also be more resilient to the deleterious effects of the GOF phenotype.

Line 423, 425, I am not a fan of the term "de-insulating!!!!!

We replaced this neologism by Similar deletion alleles resulting in a TAD fusion and misexpression effect (see below).

Line 425, why bring the work on Notch while there are so many examples in the BX-C itself....

Our revised sentence makes it more clear we are referring here to documented examples of deletion-mediated TAD fusion (ie. featuring a conformation capture assay such as HiC/micro-C):

This suggests a possible loss of the TAD boundary in the crispant clones, resulting in a TAD fusion or in a long-range interaction between a T2-specific enhancer and Ubx promoter. Similar deletion alleles resulting in a TAD fusion and misexpression effect have been described at the Notch locus in Drosophila (Arzate-Mejía et al. 2020), in digit-patterning mutants in mice and humans (Lupiáñez et al. 2015; Anania et al. 2022), or at murine and fly Hox loci depleted of CTCF-mediated regulatory blocking (Narendra et al. 2015; Gambetta and Furlong 2018; Kyrchanova et al. 2020).

Our revision also includes more emphasis on the Drosophila BX-C boundary elements Fub and Fab-7 (see above).

Version 2

Abstract

Abstract

Hox gene clusters encode transcription factors that drive regional specialization during animal development: e.g. the Hox factor Ubx is expressed in the insect metathoracic (T3) wing appendages and differentiates them from T2 mesothoracic identities. Hox transcriptional regulation requires silencing activities that prevent spurious activation and regulatory crosstalks in the wrong tissues, but this has seldom been studied in insects other than Drosophila, which shows a derived Hox dislocation into two genomic clusters that disjoined Antennapedia (Antp) and Ultrabithorax (Ubx). Here we investigated how Ubx is restricted to the hindwing in butterflies, amidst a contiguous Hox cluster. By analysing Hi-C and ATAC-seq data in the butterfly Junonia coenia, we show that a Topologically Associated Domain (TAD) maintains a hindwing-enriched profile of chromatin opening around Ubx. This TAD is bordered by a Boundary Element (BE) that separates it from a region of joined wing activity around the Antp locus. CRISPR mutational perturbation of this BE releases ectopic Ubx expression in forewings, inducing homeotic clones with hindwing identities. Further mutational interrogation of two non-coding RNA encoding regions and one putative cis-regulatory module within the Ubx TAD cause rare homeotic transformations in both directions, indicating the presence of both activating and repressing chromatin features. We also describe a series of spontaneous forewing homeotic phenotypes obtained in Heliconius butterflies, and discuss their possible mutational basis. By leveraging the extensive wing specialization found in butterflies, our initial exploration of Ubx regulation demonstrates the existence of silencing and insulating sequences that prevent its spurious expression in forewings.

Introduction

Hox genes are key specifiers of antero-posterior regional identity in animals, and thus require robust regulatory mechanisms that confine their expression to well-delimited sections of the body (Lewis 1978). Their genomic arrangement into Hox gene clusters has provided a rich template for the study of gene regulation, with mechanisms including chromatin silencing and opening, 3D conformational changes, and non-coding RNAs (Mallo and Alonso 2013). However, this rich body of work has been almost exclusively performed in mice and fruit flies. In order to decipher how diverse body plans and morphologies evolved, we must start assessing the mechanisms of Hox gene regulation in a wider range of organisms.

The Ultrabithorax (Ubx) gene encodes a Hox family transcription factor involved in the specification of segment identities in arthropods (Hughes and Kaufman 2002; Heffer and Pick 2013). In insects, the conserved expression of Ubx in the metathoracic (T3) segment is required for their differentiation from Ubx-free tissues in the mesothorax (T2), and has been a key factor for the specialization of metathoracic serial appendages including T3 legs (Mahfooz et al. 2007; Refki et al. 2014; Tomoyasu 2017; Feng et al. 2022; Buffry et al. 2023) and hindwings or their derivatives (Tomoyasu 2017; Loker et al. 2021). The mechanisms of Ubx segment-specific expression have been intensively studied in D. melanogaster (Mallo and Alonso 2013; Hajirnis and Mishra 2021), where Hox genes are separated into two genomic loci, the Antennapedia (ANT-C, Antp) and Bithorax clusters (BX-C). In short, the BX-C complex that includes Ubx, abdominal-A (abd-A), and Abdominal-B (Abd-B) is compartmentalized into nine chromosomal domains that determine the parasegmental expression boundaries of these three genes (Maeda and Karch 2015). For instance, the deletion of a small region situated between Ubx and abd-A produces the Front-ultraabdominal phenotype (Fub) where the first abdominal segment (A1) is transformed into a copy of the second abdominal segment A2, due to a gain-of-expression of abd-A in A1 where it is normally repressed (Bender and Lucas 2013). At the molecular level, the Fub boundary is enforced by insulating factors that separate Topologically Associating Domains (TADs) of open-chromatin, while also allowing interactions of Ubx and abd-A enhancers with their target promoters (Postika et al. 2018; Srinivasan and Mishra 2020). Likewise, the Fab-7 deletion, which removes a TAD boundary insulating abd-A and Abd–B (Moniot-Perron et al. 2023), transforms parasegment 11 into parasegment 12 due to an anterior gain-of-expression of Abd-B (Gyurkovics et al. 1990). By extrapolation, one may expect that if the Drosophila Hox locus was not dislocated into two complexes, Antp and Ubx 3D contact domains would be separated by a Boundary Element (BE), and that deletions similar with Fub and Fab-7 mutations would result in gain-of-function mutations of Ubx that could effectively transform T2 regions into T3 identities.

The BX-C locus also includes non-coding RNAs (Pease et al. 2013), some of which are processed into miRNAs known to repress Ubx and abd-A (Gummalla et al. 2012; Garaulet and Lai 2015). Fub-1/bxd long non-coding RNAs (lncRNAs) situated 5’ of Ubx are thought to participate in Ubx regulation in the PS5 (posterior T3 to anterior A1) parasegment (Petruk et al. 2006; Ibragimov et al. 2023). An intronic lncRNA dubbed lncRNA:PS4 is expressed in the PS4 parasegment (posterior T2 - anterior T3), and appears to stabilize Ubx in this region in mutant contexts (Hermann et al. 2022). Little is known about how insect Hox genes are regulated outside of Drosophila, where they co-localize into a single Hox cluster, and where Antp and Ubx thus occur in contiguous positions (Gaunt 2022; Mulhair and Holland 2022). A few Hox-related miRNAs are evolutionarily conserved across the locus in arthropods (Pace et al. 2016), and an early study in Tribolium characterized the embryonic expression of a Hox cluster non-coding transcripts (Shippy et al. 2008).

These knowledge gaps lead us to consider the use of butterflies and moths (Lepidoptera) as alternative model systems for the study of Ubx function and regulation. Lepidopteran forewings and hindwings are functionally and morphologically differentiated, and CRISPR mosaic knock-outs (mKOs) showed that Ubx is necessary for the specification of hindwing color patterns, shape, and venation (Tendolkar et al. 2021). In three species of nymphalid butterflies (Heliconius erato, Junonia coenia, and Bicyclus anynana), CRISPR-mediated loss-of-function of Ubx induces regional-specific homeotic transformations of hindwing patterns into their forewing counterpart (Matsuoka and Monteiro 2018; Tendolkar et al. 2021), reminiscent of homeotic aberrations that are sporadically observed in butterfly wings (Sibatani 1983; Nijhout and Rountree 1995). The ectopic activation of Ubx into the pupal forewing results in the gain of hindwing features, suggesting Ubx is sufficient to drive T3-like identity when expressed in T2 (Lewis et al. 1999; Tong et al. 2014). Besides its roles in adult wing differentiation, Ubx is also known to repress thoracic leg identity in transient embryonic appendages of the first abdominal segment, called pleuropods (Zheng et al. 1999; Masumoto et al. 2009; Tong et al. 2017; Tendolkar et al. 2021; Matsuoka et al. 2022).

The general organization of Hox gene clusters has been well described in Lepidoptera, but their regulation has been seldom studied. Lepidopteran genomes have accumulated divergent Hox3 copies, named Shox genes, that are required during early embryonic development but do not appear to play homeotic functions (Ferguson et al. 2014; Livraghi 2017; Mulhair et al. 2022). A lncRNA and two miRNAs were identified in the intergenic region between abd-A and Abd-B in the silkworm (Wang et al. 2017, 2019). In butterfly wings, the regulation of Ubx shows strong patterns of segment-specific regulation at two levels. First, the Ubx transcript has been consistently identified as the most differentially expressed mRNA between the two wing sets (Hanly et al. 2019; Wang et al. 2022). Second, comparison of ATAC-seq signals reveal that forewing vs. hindwing have identical open-chromatin profiles during development across the genome, except at the Ubx gene itself (Lewis and Reed 2018; van der Burg et al. 2019). Thus, the ability of the Ubx locus to be robustly activated in hindwings and repressed in forewings is likely driving most subsequent differences between these tissues. In this study, we provide an initial assessment of the regulation of the Ubx locus during butterfly wing development. To do this, we leverage genomic resources and CRISPR mutagenesis with a focus on two laboratory species belonging to the Nymphalinae sub-family, J. coenia and Vanessa cardui (Livraghi et al. 2017; Martin et al. 2020; van der Burg et al. 2020; Mazo-Vargas et al. 2022). We identify putative regulatory regions likely involved in the repression and activation of Ubx expression, and discuss the potential mechanisms restricting it to hindwings. Finally, we describe a collection of spontaneous wing homeotic mutants in Heliconius spp. and elaborate on the categories of mutations that could underlie these phenotypes by misregulating Ubx.

Results

Gene expression at the Ubx locus during wing development

We provide annotations of the Ubx genomic region in four Nymphalinae butterflies (Fig. 1A). These feature existing genomic resources for our model species J. coenia and V. cardui (van der Burg et al. 2020; Lohse et al. 2021b; Zhang et al. 2021), as well as for Aglais (Nymphalis) io (Lohse et al. 2021a). The publicly available annotations for these three species include evidence from developmental transcriptomes, and we added to this set a manual annotation of the Ubx locus from the oak leaf butterfly Kallima inachus, for which forewing vs. hindwing transcriptomes have been sequenced across a replicated developmental time series (Yang et al. 2020; Wang et al. 2022).

Annotation of the Ubx genomic interval in four butterflies of the Nymphalinae sub-family.

(A) Genomic intervals spanning Antp, Ubx, and abd-A, featuring published transcript annotations from NCBI Reference Genomes of V. cardui and A. io, and manual re-annotations of the J. coenia and K. inachus genomes using published RNAseq dataset (see Methods). Exons are shown with coding (thick) and non-coding (thin) sections. No lincRNA:Ubx-AS5’ transcripts were detected in K. inachus. (B) Expression profiling of transcripts of the Ubx region in K. inachus, based on a reanalysis of published wing RNA-seq transcriptomes (Wang et al. 2022). Expression levels are plotted as DESeq2 normalized counts plots. Pairwise Wald tests adjusted for multiple test correction each assess differential expression between forewings and hindwings. ns : non-significant ; * : p < 0.05; ** : p < 0.01 ; *** : p < 0.001.

All Nymphalinae show a similar organization of the region spanning Ubx. Interestingly, the first intron of Ubx encodes a long non-coding RNA in opposite orientation to Ubx, that we dub here lncRNA:Ubx-IT1 (abbr. Ubx-IT1), based on the recommended nomenclature (Seal et al. 2022). Orthologous versions of Ubx-IT1 are detected in most NCBI RefSeq genome annotations throughout Lepidoptera (e.g. the ncRNA NCBI:XR_960726 in Plutella xylostella), implying it is a conserved feature of the Ubx locus in this insect order. Finally, both annotations from V. cardui, A. io, and J. coenia show a long intergenic non-coding transcript starting in antisense orientation about 10-15 kb 5’ of Ubx, that we dub here lincRNA:Ubx-AS5’ (abbr. Ubx-AS5’). This transcript was neither detected in K. inachus transcriptomes nor in RNA datasets outside of the Nymphalinae sub-family, and could be specific to this lineage.

Next we reanalyzed the K. inachus wing transcriptomes (Wang et al. 2022), and profiled the expression of Ubx region transcripts during wing development (Fig. 1B). As expected from previous studies (Hanly et al. 2019; Paul et al. 2021; Merabet and Carnesecchi 2022; Wang et al. 2022), Ubx showed a strong expression bias in hindwings, spanning the larval imaginal disks to the intermediate pupal stage. Of note, Ubx is confined to the peripodial membranes of insect T2 wing appendages (Weatherbee et al. 1998, 1999; Prasad et al. 2016), which may explain residual detection in some of the forewing samples here. Ubx-IT1 was significantly enriched in hindwings compared to forewings, albeit at ∼100-fold lower count levels than Ubx in the same samples. The Hox gene Antp showed a minor forewing enrichment, confirming that while Ubx expression is robustly repressed in T2 forewing tissues, Antp expression is permitted in both T2 and T3 appendages (Matsuoka and Monteiro 2021, 2022; Paul et al. 2021). Expression of abd-A was undetected in most wing samples.

Chromatin 3D conformation reveals a Boundary Element between Antp and Ubx

Genome-wide Hi-C sequencing can be used to generate heatmaps that capture the conformation of 3D chromatin in tissues, and has been used extensively to study Drosophila Hox cluster organisation into TADs that prevent regulatory crosstalk between adjacent genes (Ibragimov et al. 2023; Moniot-Perron et al. 2023). Here we used Hi-C to assess the 3D chromatin architecture of the Hox cluster interval in the butterfly J. coenia, using existing datasets that were generated from fifth instar larval forewings (van der Burg et al. 2020; Mazo-Vargas et al. 2022). In larval forewings, the Hox chromatin conformation landscape consists of three well-delimited TADs, the first one spanning proboscipedia (pb) to Sex comb reduced (Scr), the second one around Antp, and the third one Ubx, abd-A, and Abd-B (Figs. 2 and 3A). A Boundary Element (BE), was robustly called (see Methods) at the region separating the Antp and Ubx TADs, situated in the Ubx last intron. Because TAD boundary prediction has a coarse resolution, we arbitrarily define the candidate BE region as a 15-kb interval centered in the Ubx last intron, and dub it Antp-Ubx_BE. A binding motif analysis identified 4 CTCF binding sites in a 1-kb interval within Antp-Ubx_BE, two of which were found in a tightly linked, convergent orientation (Fig. S1), which is consistent with TAD insulating role in mediating chromatin loop-extrusion (Guo et al. 2015). This concordance between Hi-C profiling and CTCF motif prediction thus suggests that Antp-Ubx_BE region might function as an insulator between regulatory domains of Antp and Ubx.

A region of hindwing-specific chromatin-opening is bordered by a TAD BE in the last intron of Ubx.

(A) Hi-C contact heatmap in fifth instar forewings of J. coenia and TAD separation scores around Ubx. A TAD boundary element (Antp-Ubx_BE) is inferred in the last intron of Ubx (vertical dotted line). (B) Differential ATAC-seq profiles, re-analyzed from a previous dataset (Mazo-Vargas et al. 2022). Top : open-chromatin profiles of forewings (FW, magenta), and hindwings (HW, green), respectively subtracted from larval head signal (purple, negative when wing signals at background-level). Bottom : subtractive ATAC-seq profile (HW-FW) revealing hindwing-enriched chromatin in the Ubx locus. Antp-Ubx_BE is in the vicinity of an isolated region of forewing-enriched opening (blue arrowhead). (C) PhastCons genomic alignment scores, with overall alignability suggesting minimal structural variation across this interval in Lepidoptera and Trichoptera.

Hindwing-enriched chromatin-opening around Ubx, and the Antp-Ubx_BE boundary, are both maintained in mid-pupal hindwings.

(A) Hi-C heatmap in J. coenia fifth instar larval forewings, and subtractive ATAC-seq profiles at this stage (hindwing-forewing), as expanded from Fig. 2 across the Hox cluster. (B) Hi-C heatmap in J. coenia mid-pupal hindwings, and subtractive ATAC-seq profiles at this stage (forewing-hindwing). Inferred TAD boundaries are shown as vertical dotted lines. Blue arrowhead : position of the Antp-Ubx_BE sgRNA.

Differential forewing vs. hindwing chromatin-opening across the Antp-Ubx interval

In flies, the Ubx/abd-A section is organized into regulatory domains that display differential activities across the antero-posterior axis, following what has been called the open-for-business model (Maeda and Karch 2015; Gaunt 2022). Here we tested if this pattern extends to butterfly species with a contiguous Hox cluster. To do this we used ATAC-seq datasets from J. coenia forewing (T2), hindwing (T3), and whole-head tissues sampled across fifth instar larval and early pupal stages, similarly to previous studies (van der Burg et al. 2020; Mazo-Vargas et al. 2022; Van Belleghem et al. 2023). These data reveal that chromatin opening along the Antp/Ubx/abd-A interval is partitioned into several regions showing a transition of T2 to T3 activity (Fig. 2B). From the anterior to posterior Hox collinear order (i.e. from Antp towards abd-A), chromatin-opening forms a block of forewing-enriched activity close to Antp and its 5’ region, to a block of activity in both forewings and hindwings that stops at the Antp-Ubx_BE. This region is consistent with the fact that Antp is expressed in both wing pairs (Fig. 1B). From Antp-Ubx_BE, the interval including Ubx and a large upstream region is strongly enriched for hindwing opening, consistently with previous studies that found it to be the only genomic region showing this pattern (Lewis and Reed 2018; van der Burg et al. 2019). Last, the region surrounding abd-A is devoid of differential open-chromatin activity between forewings and hindwings, in accordance with the exclusion of its expression from thoracic segments (Warren et al. 1994; Tong et al. 2014)

Comparison of 3D conformation and open-chromatin profiles between larval forewings and mid-pupal hindwings

The Hi-C dataset analyzed above was prepared from larval forewings, and forewings do not express Ubx (Fig. 1B). Next, we repeated our analysis on a Hi-C dataset generated in pupal hindwings instead (van der Burg et al. 2020), i.e. in a later-stage tissue expressing Ubx. We found two main differences in this contact landscape compared to the larval forewing (Fig. 3). First, the TAD spanning from proboscipedia (pb) to fushi-tarazu (ftz) faded in intensity, while in contrast, the TAD around Antp remained strongly organized. Second, Ubx lost its physical association to the abd-A and Abd-B domains, and gained a TAD boundary situated in the Ubx-AS5’ intron. It is difficult to disentangle effects from staging (larval vs. pupal) and tissues (forewing vs. hindwing) in this comparison. Specifically, these differences we observed may be due to chromatin remodeling between stages, as somewhat expected during metamorphosis (Gutierrez-Perez et al. 2019). Alternatively, it is also possible hindwing development requires Ubx to be insulated from the more posterior enhancers. These issues will require further investigation, for instance using profiling of histone marks, with pairwise forewing-hindwing comparison at single stages. Nonetheless the later hindwing sample showed a maintenance of Antp-Ubx separation. First, while Ubx formed a smaller TAD spanning its coding exons 1-2, this region conserved a domain of hindwing-enriched open-chromatin. Second, boundary prediction called two possible, tightly linked TAD limits in the Antp-Ubx_BE region, showing that the last intron of Ubx still acts as an insulating region. In conclusion, our preliminary comparison of Hox 3D conformation indicates that the Antp-Ubx_BE is relatively stable across two stages and wing serial homologs.

Mutagenic perturbation of Antp-Ubx_BE results in forewing homeosis

Next, we reasoned that the forewing-enriched ATAC-seq peak present in the inferred boundary interval (Fig. 4A) might mediate the binding of insulator proteins (Savitsky et al. 2016; Stadler et al. 2017), or act as a transcriptional silencer (Segert et al. 2021). Several genomic features support the former hypothesis. First, the only forewing-enriched ATAC-seq peak across a 150-kb region (spanning the Ubx gene and the Antp-Ubx intergenic region), coincides with the midpoint between the two tentative hicFindTADs boundary predictions inferred from HiC data (Fig. 2B). Second, during motif scans conducted across that 150-kb region we found 8 predicted binding-sites for the Drosophila CCCTC-Binding Factor (CTCF) clustered in a 5-kb region around the differentially accessible region, and none elsewhere in the last Ubx intron (Fig. 4A), suggesting the forewing-enriched ATAC-seq peak may function as a transcriptional insulator (Gambetta and Furlong 2018; Postika et al. 2018; Kyrchanova et al. 2020; Kaushal et al. 2022). Last, the two candidate CTCF binding motifs that are within the forewing-enriched ATAC-seq peak are also conserved across Lepidoptera and Trichoptera (Fig. S1), two lineages that diverged around 300 Mya (Kawahara et al. 2019; Thomas et al. 2020).

CRISPR perturbation of Antp-Ubx_BE results in FW➞HW homeoses.

(A) Antp-Ubx_BE sgRNA targeting (cyan triangle) of a FW-enriched ATAC-peak (magenta) within the Ubx last intron. (B-C) Two examples of J. coenia Antp-Ubx_BE crispants showing mosaic FW➞HW homeoses, shown in dorsal views. CL-WT : contralateral, horizontally flipped images of forewings from the same individuals. WT HW : wild type hindwings from the same individual and mutant forewing side. Both individuals show disruption of their Radial veins (R1-R5 area). The specimen shown in C displays a partial, ectopic eyespot (asterisk). (D-E) Immunofluorescent detection of the UbdA epitope (green) in fifth instar wings disks of Antp-Ubx_BE crispants, revealing ectopic antigenicity in forewings. WT forewings of similar stage, and HW from the same crispant individuals, are shown for comparison as insets. Green autofluorescence was observed in tracheal tissues.

To test this hypothesis, we used CRISPR targeted mutagenesis to perturb Antp-Ubx_BE and assess its functionality, and designed a single sgRNA in a conserved sequence within the forewing-enriched ATAC-seq (Fig. S1). Remarkably, CRISPR mutagenesis of the Antp-Ubx_BE target induced G0 mutants with homeotic transformations of their forewings into hindwings (Figs. 4B-C and S2), including identity shifts in patterns, venation, and wing shape. It is important to note that none of the resulting crispants showed hindwing effects. Thus, we can reasonably attribute forewing homeotic phenotypes to indel mutations restricted to the intronic region, without disruption of the Ubx transcript, as this would result in hindwing phenotypes (Matsuoka and Monteiro 2021; Tendolkar et al. 2021).

Homeotic clones are first visible in Antp-Ubx_BE crispants at the pupal stage, with streaks of thinner cuticle, sometimes associated with local necrosis or with suture defects in the ventral midline, in particular where leg and wing pouches adjoin (Fig. S3). When dissected out of the pupa, these mutant forewings also show streaks of more transparent wing epithelium, concomitantly with the territories of thinner cuticle above, and PCR genotyping of affected pupal wing tissue confirmed the presence of CRISPR-induced mutations at the Antp-Ubx_BE target site (Fig. S4). Color pattern homeotic clones were salient at the adult stage, with for example, clonal losses of the forewing specific white-band, and partial acquisitions of the large M1-M2 hindwing eyespot. In one specimen, an ectopic, partial M1-M2 hindwing eyespot appeared in the R5-M1 region, suggesting a perturbation of the eyespot induction process in this wing. Nymphalid forewings have five radial veins (R1-5), which provide sturdiness for flight (Wootton 1993), while hindwings have only two Radial veins. Forewing homeotic mutants showed mosaic venation defects in the Radial vein area (Fig. 4B). Finally, higher expressivity mutant forewings were smaller and rounder, reminiscent of hindwing shape.

Next, we dissected fifth instar larval wing disks from developing Antp-Ubx_BE crispants, and monitored the expression of Ubd-A (Ubx and Abd-A epitopes), normally restricted to the hindwing and only present in the forewing peripodial membrane (Weatherbee et al. 1999). Crispants showed forewing clones with strong ectopic expression of Ubd-A (Figs. 4D-E and S5). This result supports the inference that Antp-Ubx_BE forewing homeoses are due to the de-repression of Ubx in this tissue.

Mutational interrogation of lncRNA-encoding regions at the Ubx locus

We used CRISPR mutagenesis to test the function of the two lncRNA-encoding loci at the Ubx locus. Mutagenesis of the Ubx-IT1 first exon in J. coenia, and of the Ubx-IT1 promoter in V. cardui, both resulted in crispants with small homeotic phenotypes in forewings and hindwings (Figs. 5 and S6). This result contrasts with Ubx exon mKO experiments, which only generate hindwing phenotypes (Tendolkar et al. 2021). Given the scarcity of Ubx-IT1 crispants obtained (11 out of 236 adults), and the small size of the homeotic clones within them, we infer the occasional phenotypes may be due to rare alleles. Thus, rather than evidence of functionality of the Ubx-IT1 transcript, the homeotic phenotypes may rather reflect the effects of regulatory perturbation on Ubx itself, with some random mutations in this intronic region resulting in hindwing Ubx loss-of-function, and some others triggering derepression in forewings. Likewise, next we mutagenized the first exon of Ubx-AS5’, located upstream of the Ubx promoter, and obtained twelve hindwing mutants and a single forewing mutant (Fig. 6 and S7). As with Ubx-IT1 CRISPR experiments, these results may be explained by regulatory disruption of Ubx transcription, with a higher ratio of hindwing phenotypes compared to forewings linked to the proximity of the Ubx promoter. Overall, we conclude that the mutational interrogation at these loci can result in dual loss (hindwing) and gain (forewing) of Ubx function effects. Deciphering whether or when these effects affected Ubx expression via local cis-regulatory modules, impairment of lncRNA transcripts, or larger indels overlapping with Ubx exons, will require further study (see Discussion).

Rare, dual homeoses obtained from CRISPR mutagenesis of the lncRNA_Ubx-IT1 5’ region.

(A) Genomic context of the sgRNA targets (here shown in J. coenia), in the promoter and first exon of the non-coding Ubx-IT1 transcript. (B-C) Dorsal and ventral views of a J. coenia crispant displaying dual homeoses, i.e. with both FW➞HW (presumably due to Ubx gain-of-expression), and HW➞FW clones (akin to Ubx null mutations). Insets on the right describe forewing mutant clones (IT1 mKO), in apposition to CL-WT (contralateral forewings from the same individual), and WT HW (wild type hindwings from the same individual and mutant forewing side). (D) Examples of dual homeoses obtained when targeting orthologous sites in V. cardui.

Homeoses obtained from CRISPR mutagenesis of the lncRNA Ubx-AS5’ first exon.

(A) CRISPR sgRNA targets (here shown in J. coenia), in the first exon of the non-coding Ubx-AS5’ transcript. (B) A single J. coenia crispant showed a FW➞HW transformation. Insets on the right describe forewing mutant clones (AS5’ mKO), in apposition to CL-WT (contralateral forewings from the same individual), and WT HW (wild-type hindwings from the same individual and mutant forewing side). (C-D) Examples of HW➞FW homeoses obtained in J. coenia or when targeting orthologous sites in V. cardui. Scale bars: 500 μm.

Dual effects of mutagenesis in a putative Ubx cis-regulatory module

In an attempt to probe for Ubx hindwing-specific regulatory sequences, we focused on a ∼ 25kb region in the first intron of Ubx that displays an ATAC-seq signature of hindwing enrichment in open-chromatin relative to forewings, hereafter dubbed CRM11 (Fig. 7A). We subdivided this differentially accessible region into four peaks (11a, b, c and d). Targeting the ATAC-seq peaks with multiple sgRNAs spanning sub-domains 11a and 11c (UbxCRE11a2c5 in V. cardui, 11a2a3c5c6 in J. coenia), or with a single target in 11c (UbxCRE11c5 in V. cardui) yielded dual homeoses : FW➞HW and HW➞FW (Figs. 7B-D and S8). Hindwing effects were reminiscent of Ubx protein coding knockouts (Tendolkar et al. 2021), indicating that these crispant alleles with a hindwing phenotype produce Ubx loss-of-function effects. Individuals with hindwing clones were 2.75 times more common than individuals with forewings in this dataset. Similarly to the lncRNA loci perturbation experiments, dual homeoses may indicate the presence of hindwing activators and forewing repressors in the CRM11 region, with various CRISPR alleles producing a spectrum of indels and effects (see Discussion). It is noteworthy that while single-target experiments showed little lethality (55% hatching rate), dual or quadruple injection mixes resulted in low hatching rates of injected embryos (∼ 10%). Multiple targeting thus appears to induce high-rates of embryonic lethality, possibly due to chromosomal damage (Cullot et al. 2019; Zuccaro et al. 2020). Dual targeting with a2+c5 also yielded partial HW➞FW homeoses in V. cardui under the form of ectopic white eyespot foci phenotypes (Fig. 7E), as occasionally observed in Ubx null crispants (Tendolkar et al. 2021), seemingly due to hypomorphic or heterozygous allelic states.

CRISPR perturbation of Ubx CRM11 generates occasional dual homeotic phenotypes.

(A) Overview of ATAC-seq differential chromatin accessibility profiles (hindwing - head tissues, green ; forewing - head tissue, magenta) across the Ubx first exon. Several regions show differential opening between wings, one of which (CRM11), was targeted here for CRISPR perturbation (sites a2 and c5 indicate sgRNA targets). (B) Dual homeosis phenotypes obtained in V. cardui following dual-targeting of UbxCRE11a2c5, including homeoses in color patterns and scale morphology. (D) Additional example of a V. cardui UbxCRE11a2c5 crispant with a forewing phenotype (gain of hindwing hair patches, arrowheads). (E) Example of mild hindwing homeoses showing a white eyespot focus on the dorsal and ventral sides. These effects were previously shown to occur in coding Ubx CRISPR knock-out experiments (Tendolkar et al, 2021). Contralateral (CL) WT wings are shown for comparison with mutant wings (B-E). Colored dashed lines: wing veins. Scale bars: 500 μm.

Next, we focused on a single target shared between both V. cardui and J. coenia in the 11b sub-domain. A whole genome alignment between 23 lepidopteran species and 2 trichopteran species indicated that region 11b is deeply conserved, suggesting important functional constraints on its sequence (Fig. S9A-B). Mutagenesis of 11b using a single guide RNA (Ubx11b9) yielded a relatively high hatching rate (mean = 51.8 %), indicating a rare occurrence of the deleterious mutational effects observed in multiple targeting (see above). Four J. coenia crispants and two V. cardui crispants were obtained, all exclusively showing hindwing phenotype devoid of forewing effects. HW➞FW homeoses included a variety of phenotypes all seen in Ubx CDS mutants (Tendolkar et al. 2021), including transformations of the orange Discalis elements and the white band in J. coenia, and partial shifts in eyespot identity (Fig. S9C). Together the consistency in direction of transformations and the deep conservation of the 11b region suggests it may encode an enhancer necessary for the transcriptional activation of Ubx in hindwings.

A sample of spontaneous homeotic mutants in Heliconius butterflies

Homeotic shifts between forewings and hindwings can occur naturally in Lepidoptera, and have been documented as pattern aberrations in museum specimens (Sibatani 1980, 1983). As a complement to CRISPR-induced homeoses, we document here a rich sample of forewing/hindwing homeotic mutants in the genus Heliconius, systematically collected by L. E. Gilbert between 1987 and 2016 in captive stocks at UT Austin, as well as in the wild. Across these 15 spontaneous mutants, 12 show HW➞FW clones (Fig. S10), against 3 specimens with FW➞HW effects (Fig. 8). Mutant clones in this dataset were always posterior to the M2 vein. Only 2 mosaic phenotypes were found on a dorsal side, with the 13 others appearing ventrally. These homeotic mosaics show pattern shifts with complete fore/hindwing conversions of scale types, as seen for instance in the loss of gray wing coupling scales on posterior ventral forewings (Fig. 8A-B), or conversely, in their acquisition in posterior hindwings (arrowheads in Fig. S10B-D, H). Homeoses also include noticeable local changes in wing shape, particularly in hindwings (asterisks in Fig. S10). Taken together, these effects are akin to CRISPR-induced perturbations at the Ubx locus. We speculate that fore/hindwing homeotic aberrations, found in nature and captive stocks, result from mutations at the Ubx locus itself.

Mosaic forewing homeoses in Heliconius butterfly spontaneous mutants.

Wild-type and mutant sides from the same individuals are shown in each panel, with one side digitally flipped to match left-to-right orientation. A. Heliconius melpomene rosina, ventral view. Wild-caught in the Osa Peninsula (Costa Rica), October 1989. B. Heliconius cydno galanthus, ventral view (magnified inset in B’). Stock culture from Organisation for Tropical Studies station, La Selva (Costa Rica), June 1990 C. Heliconius himera, dorsal view (magnified inset in C’). Stock Culture in the butterfly farm Heliconius Butterfly Works in Mindo (Ecuador), March 2008.

Discussion

An intronic region with ATAC-seq hindwing-enrichment regulates Ubx

All CRISPR targets yielded homeotic phenotypes (Fig. 9), with two kinds of interference with Ubx expression – forewing gain-of-function effects, and hindwing loss-of-function effects – and indicating the presence of regulatory sequences (broadly defined), that repress or enhance Ubx expression in this region. It is crucial here to highlight the limitations of the method, in order to derive proper insights about the functionality of the regulatory regions we tested. In essence, butterfly CRISPR experiments generate random mutations by non-homologous end joining repair, that are usually deletions (Connahs et al. 2019; Mazo-Vargas et al. 2022; Van Belleghem et al. 2023). Ideally, regulatory CRISPR-induced alleles should require genotyping in a second (G1) generation to be properly matched to a phenotype (Mazo-Vargas et al. 2022). Possibly because of lethal effects, we failed to pass Ubx locus G0 mutations to a G1 generation for genotyping, and were thus limited here to mosaic analysis. As adult wings scales are dead cells, the genetic material building a given color phenotype is lost at this stage, but we circumvented this issue by genotyping a pupal forewing displaying an homeotic phenotype in the more efficient Antp-Ubx_BE perturbation experiment (Fig. S4). In this case, PCR amplification of a 600 bp fragment followed by Sanger sequencing recovered signatures of indel variants, with mixed chromatograms starting at the targeted sites. But in all other experiments (CRM11, IT1, and AS5’ targets), we did not genotype mutant tissues, as they were only detected in adult stages and generally with small clone sizes. Some of these clones may have been the results of large structural variants, as data from other organisms suggests that Cas9 nuclease targeting can generate larger than expected mutations that evade common genotyping techniques (Shin et al. 2017; Adikusuma et al. 2018; Kosicki et al. 2018; Cullot et al. 2019; Owens et al. 2019). Even under the assumption that such mutations are relatively rare in butterfly embryos, the fact we injected >100 embryos in each experiment makes their occurrence likely (Fig. 9), and we are unable to assign a specific genotype to the homeotic effects we obtained in CRM11, IT1 and AS5’ perturbation assays.

Summary of wing homeosis phenotypes obtained from mutational interrogation.

(A) CRISPR targets at non-coding regions across the Ubx region, here visualized in J. coenia. (B) Summary of injection and adult phenotype data obtained across CRISPR experiments. FW/HW crispants : total number of individuals with forewing or hindwing homeotic clones, regardless of the injected species. Individuals with dual homeosis are counted in both categories. Nmut/Ninj : number of crispants obtained (Nmut), over the number of injected embryos for each species. Bold: experiments with consistent effects in only one segment. See Table 1 for details.

When targeting hindwing-enriched ATAC-seq peaks within the first intron of Ubx – from CRM11 to the hindwing-enriched open-chromatin peak that coincides with the first exon of Ubx-IT1 – we obtained a mixture of hindwing and forewing phenotypes. Given the potential heterogeneity of allele sizes in these experiments, it is difficult to conclude robustly about the function of individual targets. Nonetheless, the phenotypic data and in particular the obtention of dual homeoses suggest we disrupted sequences that are necessary to Ubx activation in hindwings, as well as to its repression in forewings. Bifunctional cis-regulatory elements that can switch between enhancer and silencer roles are prevalent in Drosophila (Gisselbrecht et al. 2020; Segert et al. 2021; Pang et al. 2022). The CRM11 and IT1 targets adjoin or overlap with open-chromatin signals in both wing sets (Figs. 5A and 7A), providing circumstantial evidence that these regions might serve as bifunctional elements. Similar observations were recently made in mutational interrogation experiments of the butterfly WntA patterning gene (Mazo-Vargas et al. 2022). Alternatively, silencers and enhancers may be tightly linked and interact in close proximity to shape gene expression (Méndez-González et al. 2023), implying in our case that forewing and hindwing phenotypes are mediated by alleles spanning adjacent but distinct elements. A formal test of these mechanisms would require germline transmission and genotyping of these alleles, which was unsuccessful in our attempts at crossing Ubx cis-regulatory crispants.

In contrast with the dual effects obtained when targeting CRM11a+c (Fig. 9), CRM11b perturbation resulted in hindwing-limited effects, and may suggest that an Ubx enhancer was consistently compromised in this specific dataset. The high lethality and small size of mutant wing streaks suggest that only individuals with sparse, small mutant mitotic clones can survive to the adult stage. If this is true, CRM11 may contain pleiotropic enhancers that are vital for normal Ubx function at earlier stages, but expression-reporter studies will be required to test this.

Parsing lncRNA-encoding regions – correlation or cause?

LncRNAs are emerging as important regulators of gene expression and developmental processes (Zhang et al. 2019; Statello et al. 2021). IT1 targeting generated a majority of forewing phenotypes, suggesting perturbation of Ubx repression in the T2 segment. However, IT1 showed low expression in forewing RNAseq datasets from K. inachus, and higher expression in the hindwing (Fig. 1B), a pattern inconsistent with a repressive role of the antisense IT1 transcript on Ubx expression. It is generally challenging to disentangle the effects of transcription of a non-coding element from the potential effects of adjacent enhancers (Natoli and Andrau 2012; Pease et al. 2013). Therefore, an alternative explanation would be that the phenotypes are confounded by the overlap and proximity to open-chromatin regions, which may play cis-regulatory roles on Ubx via DNA-protein interactions, rather than via the lncRNA. If this is the case, it is possible that the targeted Ubx-IT1 site, which yielded homeoses in both directions and bears both forewing and hindwing open-chromatin, is a bifunctional cis-regulatory element that can shift regulatory activities between these tissues (Gisselbrecht et al. 2020). Targeted mutagenesis of the Ubx-AS5’ first exon mainly generated hindwing phenotypes, although with a relatively low-efficiency. Because this target is about 10 kb away from the Ubx promoter itself, it is plausible that the observed phenotypes were due to large deletions reaching the promoter region of Ubx. As mutational interrogation alone cannot discern if phenotypic effects arose from regulatory failure at the chromatin or transcript level, determining if AS5’ and IT1 are functional lncRNAs will need further examination. Of note, a systematic in-situ hybridization survey (Pease et al. 2013) showed that Drosophila embryos express an antisense transcript in its 5’ region (lncRNA:bxd), as well as within its first intron (lncRNA:PS4). It is thought that Drosophila bxd regulates Ubx, possibly by transcriptional interference or by facilitation of the Fub-1 boundary effect (Petruk et al. 2006; Ibragimov et al. 2023), while the possible regulatory roles of PS4 remain debated (Hermann et al. 2022). While these dipteran non-coding transcripts lack detectable sequence similarity with the lepidopteran IT1 and AS5’ transcripts, further comparative genomics analyses of the Ubx region across the holometabolan insect phylogeny should clarify the extent to which Hox cluster lncRNAs have been conserved or independently evolved.

A TAD boundary element likely acts as an insulator preventing Ubx forewing expression

Tight maintenance of TAD boundaries at the Hox locus is crucial for accurate segment identity and is facilitated by insulator proteins (Stadler et al. 2017; Gambetta and Furlong 2018; Ramírez et al. 2018). The Antp-Ubx_BE element we targeted is in a good position to block interactions between Antp and Ubx (Figs. 2-3). Consistent with this idea, the last intron of Ubx contains 8 CTCF binding motifs that are all clustered within 5-kb around the forewing-enriched ATAC peak, including two sites at highly conserved positions that are only 100-bp away from the CRISPR target (Fig. S1). CTCF sites prevent cross-talk between regulatory domains in the fly BX-C complex, and result in Hox misexpression when deleted (Postika et al. 2018; Kyrchanova et al. 2020; Kaushal et al. 2022; Kahn et al. 2023). Thus, the density of CTCF sites in this region may be indicative of a bona fide insulator active in forewings.

CRISPR mutagenesis of Antp-Ubx_BE generated FW➞HW homeoses associated with a gain of UbdA antigenicity in forewings, with no effects in the other direction, in stark contrast with other targets (Fig. 9B). This suggests a possible loss of the TAD boundary in the crispant clones, resulting in a TAD fusion or in a long-range interaction between a T2-specific enhancer and Ubx promoter. Similar deletion alleles resulting in a TAD fusion and misexpression effect have been described at the Notch locus in Drosophila (Arzate-Mejía et al. 2020), in digit-patterning mutants in mice and humans (Lupiáñez et al. 2015; Anania et al. 2022), or at murine and fly Hox loci depleted of CTCF-mediated regulatory blocking (Narendra et al. 2015; Gambetta and Furlong 2018; Kyrchanova et al. 2020). It will be interesting to profile the binding of insulator proteins and transcriptional repressors across the butterfly Hox TAD landscape to shed more light onto the mechanisms of Ubx insulation, using in vivo assays (Bowman et al. 2014), or in silico predictions that take advantage of updated binding matrices for insect insulator proteins (Mitra et al. 2018). Of note, our CRISPR target is adjacent to an hindwing-enriched peak that also presented CTCF binding sites (Fig. 4A). Following a similar logic, this site could be a candidate insulator specific to Ubx-expressing tissues like the hindwing, a hypothesis that will require further testing. Lastly, it is worth noting that the Antp/Ubx TAD boundary we identified is intragenic, within the last intron of Ubx. It is unclear if this feature affects transcription, but this configuration might be analogue to the Notch locus in Drosophila, which includes a functional TAD boundary in an intronic position (Arzate-Mejía et al. 2020).

Making sense of spontaneous wing homeotic mutants

In this article, we documented a large sample of spontaneous homeotic mutants obtained in Heliconius spp. All homeotic clones were limited to the wing posterior compartments (ie. posterior to the M2 vein), possibly because of parasegmental, compartment-specific regulatory domains that played historic roles in the study of Drosophila BX-C regulation (Maeda and Karch 2015). Sibatani documented in Lepidoptera that “the mosaics of F/H homeosis tend to occur most frequently in the posterior half of the wing, the boundary of the anterior and posterior halves occurring somewhere in space M1-M2” (Sibatani 1983). Our collection of spontaneous Heliconius mutants only displayed clones in posterior regions, consistently with this trend. However, our CRISPR perturbation assays of J. coenia and V. cardui cis-regulatory regions also generated anterior clones, with all targets. Deciphering how butterfly Ubx regulation is compartmentized between parasegmental or wing antero-posterior domains will require additional investigation. Most Heliconius homeoses were in the hindwings (ie. putative Ubx loss-of-expression clones), and among these, all but one were ventral (Fig. S10). Three mutants showed forewing homeoses (ie. putative Ubx gain-of-expression clones), two of them ventral and one of them dorsal (Fig. 8). The systematic reviews of wing homeosis in Lepidoptera found that forewing homeoses are almost as common as hindwing ones (Sibatani 1980, 1983). Our mutational interrogation assays, while coarse in nature, revealed the existence of activating and repressing cis-regulatory sequences at the Ubx locus itself. Spontaneous FW↔HW homeoses observed in butterflies and moths may thus result from somatic mutations or transposition events at this locus.

Materials and methods

Genome annotations and transcriptomic analysis

Nymphalid genome sequences of the Hox cluster and their annotations were extracted from the NCBI Assembly and Lepbase online repositories (Challis et al. 2016; Kitts et al. 2016) as follows : V. cardui from NCBI ilVanCard2.1 and LepBase Vc_v1 ; A. (Nymphalis) io from NCBI ilAglIoxx1.1; J. coenia from Lepbase Jc_v2; P xylostella from NCBI ilPluXylo3.1. The Ubx regions from ilVanCard2.2, Vc_v1, and Jc_v2 were manually re-annotated using wing transcriptome data on the NCBI SRA (BioProjects PRJNA661999, PRJNA293289, PRJNA237755, PRJNA385867, and PRJNA498283) The genome sequence of K. inachus was obtained from the Dryad repository (Yang et al. 2020). Differential gene expression analysis across the K. inachus Ubx locus were carried out using wing transcriptome data available on the NCBI SRA (BioProject PRJNA698433), following a manual re-annotation of a preliminary gene models provided by Peiwen Yang and Wei Zhang (Wang et al. 2022). All transcripts analyses were performed using the STAR intron-aware aligner and DEseq2 expression analysis package as previously described (Love et al. 2014; Dobin and Gingeras 2016; Hanly et al. 2019, 2022). Expression levels were calculated as genome-wide normalized counts and pairwise Wald tests were performed to assess differential expression between forewings and hindwings. Multiple test adjustment was performed using Benjamini and Hochberg correction.

Hi-C and ATAC-seq analyses

Hi-C data from J. coenia fifth instar larval forewings and 48-72 h APF pupa hindwings are available at the NCBI SRA BioProject PRJNA641138 (van der Burg et al. 2020). Triplicated ATAC-seq datasets for larval and pupal wing and head tissues of J. coenia and V. cardui (van der Burg et al. 2019, 2020; Mazo-Vargas et al. 2022) are available on the NCBI SRA BioProjects PRJNA497878, PRJNA695303, and PRJNA559165. All the ATAC-seq and Hi-C data were re-analysed on J. coenia and V. cardui Ubx genomic regions as previously described (Mazo-Vargas et al. 2022). Briefly, matrices of interactions were constructed by mapping paired reads against the Junonia coenia genome (Mazo-Vargas et al., 2022) using hicBuildMatrix (Ramírez et al. 2018). Matrices from larvae and pupae were normalized using hicNormalize and corrected with the Knight-Ruiz matrix balancing algorithm. The definitions of topologically associating domains (TADs) can be influenced by various factors such as the choice of software, parameters, sequencing depth, and the presence of experimental noise. To ensure reliability, it is recommended to compare TAD calls with independent datasets, such as histone marks or known factors associated with TAD boundaries. In the absence of these specific datasets, we employed a different combination of parameters in the hicFindTADs tool and compared the resulting TAD calls. HiC matrices at 10 kb and 20 kb bin resolutions were utilized, and TAD insulation scores were evaluated using a false-discovery rate correction for multiple testing, with p-value thresholds of 0.01 and 0.005. Consistent TAD boundary calls with negative TAD separation scores were selected to define domain limits at 10 kb and 20 kb matrix resolutions.

CTCF motif binding predictions

The program fimo was used to scan for the J. coenia candidate TAD boundary region (HiC_scaffold_12:6430000-6444000) for canonical CTCF binding sites, using the positional weight matrix MA0205.1 deposited in the JASPAR database (Holohan et al. 2007; Cuellar-Partida et al. 2012; Castro-Mondragon et al. 2022).

Genomic conservation analyses

We generated whole-genome alignments of 25 Lepidoptera and 2 Trichoptera reference species from NCBI Assembly using ProgressiveCactus (Armstrong et al. 2020), and HAL tools (Hickey et al. 2013) for converting the resulting HAL file to the MAF format. We provided a species topology tree of 23 Lepidoptera species to PhyloFit (Hubisz et al. 2011) to fit a multiple sequence alignment on the reference J. coenia Ubx locus, using HKY85 as the substitution model. Using PhastCons (Siepel et al. 2005), we then generated conservation score plots using standard parameters (target-coverage = 0.45; expected-length = 12; rho = 0.1) stored in BED and WIG file formats.

Butterfly rearing and CRISPR microinjections

J. coenia and V. cardui colonies were maintained at 25°C and 60-70% relative humidity in a growth chamber with a 14:10 light:dark photoperiod. Larval rearing on artificial diets, egg collection, and microinjections followed previously described methods (Martin et al. 2020; Tendolkar et al. 2021). Cas9:sgRNA heteroduplexes were prepared as previously described (Martin et al., 2020). Frozen aliquots of Cas9-2xNLS (2.5 μL ; 1,000 ng/μL) and sgRNA (2.5 μL ; 500 ng/μL) were mixed in 2:1 and 4:1:1 mass ratios for single and dual target injections, respectively. CRISPR sgRNA targets are listed in Table S1.

Genotyping

For verification in DNA lesions at the intended Antp-Ubx_BE site, a pupal wing fragment harboring visible mutant clones (Fig. S4B”) or control wild-type tissue were PCR amplified using the diluted protocol of the Phire Animal Tissue Direct PCR Kit (ThermoFisher) and a pair of oligonucleotides (Forward : 5’-ACCGATCGTAAACGTCAACTTAACG-3’ ; Reverse 5’-TACTGCGGTGGCGAGTGAATG-3’), before purification and Sanger sequencing using the reverse primer.

Antibody stainings

Fifth instar wing disks were dissected in ice cold Phosphate Buffer Saline (PBS), fixed for 15-20 min at room temperature in methanol-free formaldehyde diluted to 4% in PBS / 2 mM EGTA (egtazic acid), washed in PBS with 0.1% Triton X-100 (PT), stored in PT with 0.5% Bovine Serum albumin (PT-BSA), incubated overnight at 4°C in PT-BSA with a 1:5 dilution of the anti-UbdA peptide antibody serum (mouse monoclonal FP6.87, Developmental Studies Hybridoma Bank), and washed in PT. A 1:250 dilution of anti-Mouse IgG antibody coupled to AlexaFluor488 or Rabbit AlexaFluor555 was made in PT-BSA and spun down at 14,000 rcf to remove aggregates, and incubated with wings for 2 h at room temperature, before additional washes, incubation in 50% glycerol-PBS with DAPI (4′,6-diamidino-2-phenylindole) nuclear stain, and incubation and mounting in 60% glycerol-PBS with 2 mM of EDTA (Ethylenediaminetetraacetic acid).

Imaging

Full-mount photographs of J. coenia and V. cardui were taken on a Nikon D5300 digital camera mounted with an AF-S VR MicroNikkor 105mm f/2.8G lens, with magnified views taken using a Keyence VHX-5000 digital microscope mounted with VH-Z00T and VH-Z100T lenses. Immunofluorescent stainings were imaged on an Olympus BX53 epifluorescent microscope mounted with UPLFLN 4x/0.13 and 10X/0.3 objectives.

Supporting information